---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---


```{r, include=(FALSE)}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=TRUE, results="verbatim")
options(knitr.duplicate.label = "allow")

library(Seurat)
library(WGCNA) # pour la fonction collapseRows
library(edgeR)
```

# Parameters

Here we set up input and outputs files for the vignette.

```{r}
if (!exists("reference_file")) {reference_file = "01_references/reference_pdac.rds"} 

input_dir = "~/projects/datashare/input_ref_hadaca3/" # on lucie's gricad dahu
```

# Generate

## RNA

Needed files : 

GENEANNOT_human_rnannot_complete.RDS
DATA_PDX/Human-Tumor_rawcount_Transcriptome.tsv

Dir: 
GSE135123_RAW_transcripto
GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_reads_LL.tsv
GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt

```{r}
### GENE ANNOTATIONS ###
annot_genes <- readRDS(paste0(input_dir,"GENEANNOT_human_rnannot_complete.RDS"))

### PDX DATA  https://github.com/RemyNicolle/PaCaOmicsDATA  ###
#Selection of samples of interest 
ID_basalpurePDX <- c("PDAC001T","PDAC004T","PDAC028T", "PDAC024T")
ID_classicalpurePDX <- c("PDAC018T","PDAC011T","PDAC026T","PDAC006T","PDAC030T")

#Load expression data
PDX_RNASEQ_BATCH1_RAWCOUNTS <- read.delim(paste0(input_dir,"DATA_PDX/Human-Tumor_rawcount_Transcriptome.tsv"), row.names = 1)

#From ENSG ID to SYMBOLS 
PDX_RNASEQ_BATCH1_RAWCOUNTS_GENESYMBOL <- collapseRows(PDX_RNASEQ_BATCH1_RAWCOUNTS, annot_genes[rownames(PDX_RNASEQ_BATCH1_RAWCOUNTS),"GeneName"],  rownames(PDX_RNASEQ_BATCH1_RAWCOUNTS), method="MaxMean")$datET

DATA_PDX_transcriptome <- data.frame(TUM_BASAL= PDX_RNASEQ_BATCH1_RAWCOUNTS_GENESYMBOL[,ID_basalpurePDX], TUM_CLASSIC = PDX_RNASEQ_BATCH1_RAWCOUNTS_GENESYMBOL[,ID_classicalpurePDX])
#saveRDS(DATA_PDX_transcriptome, file = "DATA_PDX_transcriptome.rds")
DATA.PDX <- DATA_PDX_transcriptome


### ENDOTHELIAL CELL DATA  ###
files = list.files(path = paste0(input_dir,"GSE135123_RAW_transcripto"), pattern=".*txt")
endo_GSE23 = lapply(files, function(file) {
  counts = read.delim( paste0(input_dir,"GSE135123_RAW_transcripto/",file), sep = "\t", header = F)
  colnames(counts) = c("Gene",paste0("ENDO.",substr(file,1,10)))
  return(counts)
} )
endo_df_GSE23 = do.call(cbind, lapply(endo_GSE23, function(df) df[ , 2, drop = FALSE])) #combine all counts 
rownames(endo_df_GSE23) = endo_GSE23[[1]]$Gene
DATA.GSE23 =  collapseRows(endo_df_GSE23, annot_genes[rownames(endo_df_GSE23),"GeneName"],  rownames(endo_df_GSE23),method="MaxMean")$datET


### GETEX DATA ###
datagetex = read.delim(paste0(input_dir,"GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_reads_LL.tsv"),sep="\t",header = T,row.names=1) 
# to bypass the use of read.gct(couldn't install CePa package) I used "sed '1,2d' GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_reads.gct > GTEx_Analysis_2017-06-05_v8_RNASeQCv1.1.9_gene_reads_LL.tsv" to remove the first 2 rows of the file

colnames(datagetex)=gsub("[.]", "",colnames(datagetex))

annotgetex =read.delim(paste0(input_dir,"GTEx_Analysis_v8_Annotations_SampleAttributesDS.txt"),header=T, sep="\t",row.names=1)
rownames(annotgetex)=gsub("-", "",rownames(annotgetex))

### Selection of tissues of interest ###
fibro = rownames(annotgetex)[which(annotgetex$SMTSD=="Cells - Cultured fibroblasts")]
spleen = rownames(annotgetex)[which(annotgetex$SMTSD=="Spleen")] # cellules immunitaires 
fibroID = intersect(colnames(datagetex),fibro )
spleenID = intersect(colnames(datagetex),spleen )
DATA.GETEX = data.frame(FIBRO = datagetex[,fibroID], IMMUNE = datagetex[,spleenID])

### From ENSG ID to SYMBOLS ###
namesID =unlist(lapply(strsplit(rownames(DATA.GETEX), "[.]"), function(x)x[1]))
DATA.GETEXENSG = collapseRows(DATA.GETEX, namesID,  rownames(DATA.GETEX),method="MaxMean")$datET
DATA.GETEXrawg  = collapseRows(DATA.GETEXENSG, annot_genes[rownames(DATA.GETEXENSG),"GeneName"],  rownames(DATA.GETEXENSG),method="MaxMean")$datET
#saveRDS(DATA.GETEXrawg, "DATA.GETEXrawg.RDS")


### MERGE DATA ###
commongenes <- Reduce(intersect, list(rownames(DATA.GETEXrawg),rownames(DATA.PDX) ,rownames(DATA.GSE23))) #intersects all of the genes
DATALLraw = data.frame(DATA.GETEXrawg[commongenes,],DATA.PDX[commongenes,],DATA.GSE23[commongenes,])
#> dim(DATALLraw)
#[1] 27786   777
cl = as.factor(unlist(lapply(strsplit(colnames(DATALLraw),"[.]"), function(x)x[1]))) # get all CT from dataset colnames
series = as.factor(c(rep("GTEX", ncol(DATA.GETEXrawg)), rep("PDX", ncol(DATA.PDX)),rep("GSE23", ncol(DATA.GSE23))))
annotdata = data.frame(Population=cl, Series = series)
rownames(annotdata) = colnames(DATALLraw)
Data_DC_transcriptomeRAW = list(expr = DATALLraw, annot = annotdata)
#save(Data_DC_transcriptomeRAW, file=paste0(input_dir,"Data_DC_transcriptomeRAW_LL.RData"))
write.table(DATALLraw, "01_references/ref_bulkRNA_raw_full.txt", sep = "\t")


### AGGREGATE COLUMNS RAW DATA ###

idx_fib = which(substring(colnames(DATALLraw),1,5) == "FIBRO")
t_fib = round(apply(DATALLraw[, idx_fib], 1, median),0)
idx_immune = which(substring(colnames(DATALLraw),1,6) == "IMMUNE")
t_immune = round(apply(DATALLraw[, idx_immune], 1, median),0)
idx_classic = which(substring(colnames(DATALLraw),1,11) == "TUM_CLASSIC")
t_classic = round(apply(DATALLraw[, idx_classic], 1, median),0)
idx_basal = which(substring(colnames(DATALLraw),1,9) == "TUM_BASAL")
t_basal = round(apply(DATALLraw[, idx_basal], 1, median),0)
idx_endo = which(substring(colnames(DATALLraw),1,4) == "ENDO")
t_endo = round(apply(DATALLraw[, idx_endo], 1, median),0)

ref_rna_raw_agreggated = cbind(endo = t_endo,
                  fibro = t_fib,
                  immune = t_immune,
                  classic = t_classic,
                  basal = t_basal)
write.table(ref_rna_raw_agreggated, "01_references/ref_bulkRNA_raw.txt", sep = "\t")



### NORMALIZE AND AGGREGATE COLUMNS ###

dge = DGEList(counts=DATALLraw) # create DGEList object
dge = calcNormFactors(dge) #Calculate Normalization Factors to Align Columns of a Count Matrix
DATAcpm = cpm(dge, normalized.lib.sizes=TRUE, log=TRUE) #normalized count less than 1 will be negative # Computes counts per million (cpm)

DATAcpmnotlog = cpm(dge,normalized.lib.sizes=TRUE, log=FALSE)
DATAcpmlog2 = log2(DATAcpmnotlog + 1) #ensure non negative values

idx_fib = which(substring(colnames(DATAcpmnotlog),1,5) == "FIBRO")
t_fib = apply(DATAcpmnotlog[, idx_fib], 1, median)
idx_immune = which(substring(colnames(DATAcpmnotlog),1,6) == "IMMUNE")
t_immune = apply(DATAcpmnotlog[, idx_immune], 1, median)
idx_classic = which(substring(colnames(DATAcpmnotlog),1,11) == "TUM_CLASSIC")
t_classic = apply(DATAcpmnotlog[, idx_classic], 1, median)
idx_basal = which(substring(colnames(DATAcpmnotlog),1,9) == "TUM_BASAL")
t_basal = apply(DATAcpmnotlog[, idx_basal], 1, median)
idx_endo = which(substring(colnames(DATAcpmnotlog),1,4) == "ENDO")
t_endo = apply(DATAcpmnotlog[, idx_endo], 1, median)

ref_rna = cbind(endo = t_endo,
                  fibro = t_fib, 
                  immune = t_immune, 
                  classic = t_classic,
                  basal = t_basal)

write.table(ref_rna, "01_references/ref_bulkRNA.txt", sep = "\t")

```

## Methylation

Needed files : 
Data_DC_methylome.RData
DATA_PDX_methylome.RDS
beta value files from GEO GSE82234 

Generated ref for hadaca3 based on hadaca2, with full intermediary sample results. Only beta value previously computed (either by yuna or directly downloaded from geo). Raw value would be to heavy to compute and load. 
```{r}
#Load yuna's Rdata from hadaca2 reference generation, is a mix of PDX (unpublished for some), CAF et TRIS 
load(paste0(input_dir,"Data_DC_methylome.RData"))
h2_ref = Data_DC_methylome$meth

#Load yuna's Rdata, published pdx data 
pdx_pub = readRDS(paste0(input_dir,"DATA_PDX/DATA_PDX_methylome.RDS"))

#Merging yuna's rdata ouput from reference generation for hadaca2 with endo cells
# Load data from GEO GSE82234 

files = list.files(path = paste0(input_dir,"GSE82234_RAW_methylo/"), pattern=".*tsv")
endo = lapply(files, function(file) {
  met = read.delim( paste0(input_dir,"GSE82234_RAW_methylo/",file), sep = "\t", header = T)
  colnames(met) = c("Probe",paste0("ENDO.",substr(file,1,10)))
  return(met)
} )
endo_df = do.call(cbind, lapply(endo, function(df) df[ , 2, drop = FALSE])) #combine all counts
rownames(endo_df) = endo[[1]]$Probe

na_rows = which(is.na(endo_df), arr.ind = TRUE)[,1]
endo_df = endo_df[-na_rows,]


commonprobes = intersect(rownames(endo_df), rownames(h2_ref))
commonprobes = intersect(commonprobes, rownames(pdx_pub))


#remove tum basal an tum classic from h2 because we will be using only published pdx
tris_caf = c("FIBRO_C1","FIBRO_C2","FIBRO_C3","FIBRO_C4","FIBRO_N1","FIBRO_N2","FIBRO_N3","FIBRO_N4","IMMUNE_HACK02CFPNID1r1","IMMUNE_HACK02CFPNID1r2","IMMUNE_HACK02CFPNID1r3")

datameth_all = data.frame(h2_ref[commonprobes,tris_caf],pdx_pub[commonprobes,],endo_df[commonprobes,])

colnames(datameth_all)
rename_vec = c(
"FIBRO_C1"="FIBRO_CANCER_1",
"FIBRO_C2"="FIBRO_CANCER_2",
"FIBRO_C3"="FIBRO_CANCER_3",
"FIBRO_C4"="FIBRO_CANCER_4",
"FIBRO_N1"="FIBRO_NORMAL_1",
"FIBRO_N2"="FIBRO_NORMAL_2",
"FIBRO_N3"="FIBRO_NORMAL_3",
"FIBRO_N4"="FIBRO_NORMAL_4",
"IMMUNE_HACK02CFPNID1r1"="IMMUNE_1",
"IMMUNE_HACK02CFPNID1r2"="IMMUNE_2",
"IMMUNE_HACK02CFPNID1r3"="IMMUNE_3",
"TUM_BASAL.PDAC004T"="TUMOR_BASAL_1",
"TUM_BASAL.PDAC001T"="TUMOR_BASAL_2",
"TUM_BASAL.PDAC028T"="TUMOR_BASAL_3",
"TUM_BASAL.PDAC024T"="TUMOR_BASAL_4",
"TUM_CLASSIC.PDAC018T"="TUMOR_CLASSIC_1",
"TUM_CLASSIC.PDAC011T"="TUMOR_CLASSIC_2",
"TUM_CLASSIC.PDAC006T"="TUMOR_CLASSIC_3",
"TUM_CLASSIC.PDAC030T"="TUMOR_CLASSIC_4",
"TUM_CLASSIC.PDAC026T"="TUMOR_CLASSIC_5",
"ENDO.GSM2186934"="ENDO_1",
"ENDO.GSM2186935"="ENDO_2",
"ENDO.GSM2186936"="ENDO_3",
"ENDO.GSM2186937"="ENDO_4",
"ENDO.GSM2186938"="ENDO_5",
"ENDO.GSM2186939"="ENDO_6" )

colnames(datameth_all)[match(names(rename_vec), colnames(datameth_all))] = rename_vec

write.table(datameth_all, "01_references/ref_met_betaval_full.txt", sep = "\t")

#from h2_ref 
idx_fib = which(substring(colnames(datameth_all),1,5) == "FIBRO")
t_fib = apply(datameth_all[, idx_fib], 1, median)
idx_immune = which(substring(colnames(datameth_all),1,6) == "IMMUNE")
t_immune = apply(datameth_all[, idx_immune], 1, median)

#from pdx_pub
idx_classic = which(substring(colnames(datameth_all),1,14) == "TUMOR_CLASSIC_")
t_classic = apply(datameth_all[, idx_classic], 1, median)
idx_basal = which(substring(colnames(datameth_all),1,12) == "TUMOR_BASAL_")
t_basal = apply(datameth_all[, idx_basal], 1, median)

# from geo endo 
idx_endo = which(substring(colnames(datameth_all),1,4) == "ENDO")
t_endo = apply(datameth_all[, idx_endo], 1, median)


ref_met = cbind(endo = t_endo,
                fibro = t_fib, 
                immune = t_immune, 
                classic = t_classic,
                basal = t_basal)

write.table(ref_met, "01_references/ref_met_betaval.txt", sep = "\t")
```

## Single cell RNAseq

### Peng 

```{r}
peng = readRDS(paste0(input_dir, "00_peng_k_2019.rds"))

# get all indices 
idx_endo <- which(peng@meta.data$cell_type == "Endothelial cell")
idx_immune <- which(peng@meta.data$cell_type %in% c("B cell","T cell","Macrophage cell"))
idx_fibro <- which(peng@meta.data$cell_type == c("Stellate cell","Fibroblast cell"))
idx_tumor <- which(peng@meta.data$cell_type == "Ductal cell type 2")


peng@meta.data[idx_endo,"hadaca3"] <- "endo" 
peng@meta.data[idx_immune,"hadaca3"] <- "immune"
peng@meta.data[idx_fibro,"hadaca3"] <- "fibro"
peng@meta.data[idx_tumor,"hadaca3"] <- "tumor"

peng = subset(peng, subset = hadaca3 %in% c("endo","immune","fibro","tumor"))

sampled_cells <- sample(Cells(peng), 5000)
peng_sub = subset(peng, cells = sampled_cells)
```

### Baron 

```{r}
baron = readRDS(paste0(input_dir,"00_baron_m_2016.rds"))

# get all indices 
idx_endo <- which(baron@meta.data$cell_type == "endothelial")
idx_immune <- which(baron@meta.data$cell_type %in% c("macrophage","mast","t_cell"))
idx_fibro <- which(baron@meta.data$cell_type == c("activated_stellate", "quiescent_stellate"))

baron@meta.data[idx_endo,"hadaca3"] <- "endo" 
baron@meta.data[idx_immune,"hadaca3"] <- "immune"
baron@meta.data[idx_fibro,"hadaca3"] <- "fibro"

baron_sub = subset(baron, subset = hadaca3 %in% c("endo","immune","fibro"))
baron_sub$hadaca3 = as.factor(baron_sub$hadaca3)
#no sub_set necessary only 551 cell remaining 

counts_baron =  baron_sub@assays$RNA$counts #type dgCMatrix (sparse matrix)
metadata_baron = data.frame(cell_type = baron_sub@meta.data$hadaca3, sample = baron_sub@meta.data$sample, tissue = baron_sub@meta.data$tissue)
rownames(metadata_baron) <- rownames(baron_sub@meta.data)
```

### Raghavan 

```{r}
raghavan = readRDS(paste0(input_dir,"00_raghavan_s_2021.rds"))

# get all indices 
idx_endo <- which(raghavan@meta.data$cell_type == "Endothelial")
idx_immune <- which(raghavan@meta.data$cell_type %in% c("B_Cells","Macrophage","T_NK","T_Regs","DC","pDC_cell"))
idx_tumor <- which(raghavan@meta.data$cell_type == "Tumor")

raghavan@meta.data[idx_endo,"hadaca3"] <- "endo" 
raghavan@meta.data[idx_immune,"hadaca3"] <- "immune"
raghavan@meta.data[idx_tumor,"hadaca3"] <- "tumor"


raghavan = subset(raghavan, subset = hadaca3 %in% c("endo","immune","tumor"))
sampled_cells <- sample(Cells(raghavan), 5000)
raghavan_sub = subset(raghavan, cells =  sampled_cells)
```

```{r}
rm(raghavan)
rm(baron)
rm(peng)
gc()
```

### Label classic and basal 

#### Run puriST

```{r}
source("01-additional/functions_puriST.R")
load("01-additional/fitteds_public_2019-02-12_puriST.Rdata") # load classifiers

classifier = classifs[[1]]

# peng
peng_t = subset(peng_sub, subset = hadaca3 == "tumor")

peng_t = SCTransform(peng_t)

predictions_peng = apply_classifier(data = peng_t@assays$RNA$counts, classifier = classifier)

table(predictions_peng$Subtype)
table(predictions_peng$Subtype_graded)

#replace names to fit 
predictions_peng$Subtype[predictions_peng$Subtype_graded %in% c("likely basal-like","strong basal-like")] <- "basal"
predictions_peng$Subtype[predictions_peng$Subtype_graded == "lean basal-like"] <- "undefined_tumor"

predictions_peng$Subtype[predictions_peng$Subtype_graded  %in% c("likely classical","strong classical")] <- "classic"
predictions_peng$Subtype[predictions_peng$Subtype_graded  == "lean classical"] <-  "undefined_tumor"

peng_t$purist = predictions_peng$Subtype

#update main pool
peng_sub$hadaca3[names(peng_t$purist)] = as.character(peng_t$purist)

 # remove undefined tumor cells 
keep_cell <- rownames(peng_sub@meta.data[peng_sub@meta.data$hadaca3 != "undefined_tumor", ])

peng_sub = subset(peng_sub, cells =  keep_cell)

counts_peng =  peng_sub@assays$RNA$counts #type dgCMatrix (sparse matrix)
metadata_peng = data.frame(cell_type = peng_sub@meta.data$hadaca3, sample = peng_sub@meta.data$sample, tissue = peng_sub@meta.data$tissue)
rownames(metadata_peng) <- rownames(peng_sub@meta.data)
```

```{r}
#raghavan
raghavan_t = subset(raghavan_sub, subset = hadaca3 == "tumor")

raghavan_t = SCTransform(raghavan_t)

predictions_raghavan = apply_classifier(data = raghavan_t@assays$RNA$counts, classifier = classifier)

table(predictions_raghavan$Subtype)

#replace names to fit 
predictions_raghavan$Subtype[predictions_raghavan$Subtype_graded %in% c("likely basal-like","strong basal-like")] <- "basal"
predictions_raghavan$Subtype[predictions_raghavan$Subtype_graded == "lean basal-like"] <- "undefined_tumor"

predictions_raghavan$Subtype[predictions_raghavan$Subtype_graded  %in% c("likely classical","strong classical")] <- "classic"
predictions_raghavan$Subtype[predictions_raghavan$Subtype_graded  == "lean classical"] <-  "undefined_tumor"

raghavan_t$purist = predictions_raghavan$Subtype
#update main pool
raghavan_sub$hadaca3[names(raghavan_t$purist)] = as.character(raghavan_t$purist)
 # remove undefined tumor cells 
keep_cell <- rownames(raghavan_sub@meta.data[raghavan_sub@meta.data$hadaca3 != "undefined_tumor", ])

raghavan_sub = subset(raghavan_sub, cells =  keep_cell)

counts_raghavan =  raghavan_sub@assays$RNA$counts #type dgCMatrix (sparse matrix)
metadata_raghavan = data.frame(cell_type = raghavan_sub@meta.data$hadaca3, sample = raghavan_sub@meta.data$sample,tissue = raghavan_sub@meta.data$tissue)
rownames(metadata_raghavan) <- rownames(raghavan_sub@meta.data)
```

### Plot 

```{r, eval=FALSE}
# if (!requireNamespace("BiocManager", quietly = TRUE))
#     install.packages("BiocManager")
# BiocManager::install("ComplexHeatmap")
library(ComplexHeatmap)
```

```{r, eval=FALSE}
zs = function(mtx) {
  t(apply(mtx, 1, function(x) (x - mean(x)) / sd(x)))
}

plot_heatmap = function (df, markers, fig_name){
  
  idx_to_keep = which(markers$Gene %in% rownames(df))
  markers = markers[idx_to_keep, ]
  heatmap_val = zs(df[markers$Gene,])
   x = apply(heatmap_val, 1, function(r){
    sum(is.na(r))
  })
  idx_na_rows = which(x>0)
  if (length(idx_na_rows)> 0) {
    heatmap_val  =heatmap_val [-idx_na_rows, ]
  }
  # dim(heatmap_val)
  cur_markers = markers[markers$Gene %in% rownames(heatmap_val), ]
  cur_markers$short = substring(cur_markers$Compartment, 1, 3)
  heatmap_val = heatmap_val[cur_markers$Gene,]
  
  hm = Heatmap(heatmap_val,
               #bottom_annotation = column_ha1,
               row_split = cur_markers$short,
               show_row_names=F,
               column_names_gp = gpar(fontsize = 12),
               row_names_gp = gpar(fontsize = 12), 
               name = "Z-score",
               
               column_title  = fig_name)
  return(hm)
}
```

```{r, eval=FALSE}
rob_markers = c("TFF1", "GATA6", "HNF1A", "CDN18", "S100A2", "MET", "EGFR", "MUC16", "CYP24A1")
rob_df = data.frame(Compartment = c("Classic", "Classic", "Classic", "Classic", "Basal", "Basal", "Basal", "Basal", "Basal"),
                    Gene = rob_markers)

length(intersect(rob_markers,rownames(peng_sub@assays$RNA$counts)))
length(intersect(rob_markers,rownames(raghavan_sub@assays$RNA$counts)))

setdiff(rob_markers,rownames(peng_sub@assays$RNA$counts))
setdiff(rob_markers,rownames(raghavan_sub@assays$RNA$counts))

length(rob_markers)

#remove missing gene CDN18
rob_markers = c("TFF1", "GATA6", "HNF1A", "S100A2", "MET", "EGFR", "MUC16", "CYP24A1")
rob_df = data.frame(Compartment = c("Classic", "Classic", "Classic", "Basal", "Basal", "Basal", "Basal", "Basal"),
                    Gene = rob_markers)

```

```{r, eval=FALSE}
# RAGHAVAN
raghavan_t = subset(raghavan_sub, subset = hadaca3 %in% c("classic","basal"))

#subset to 100 basal and 100 classic 
basal = subset(raghavan_t, hadaca3 == "basal")
classic = subset(raghavan_t, hadaca3 == "classic")

sample_cl <- sample(Cells(classic), 100)

classic = subset(classic, cells =  sample_cl)

subset = merge(classic,basal)
subset = JoinLayers(subset)

df = subset@assays$RNA$counts
colnames(df) = subset$hadaca3

# purist markers 
markers = data.frame(Compartment = c(rep("Classic",21),rep("Basal",21)),
                    Gene = c(classifier$TSPs[,2],classifier$TSPs[,1]))



plot_heatmap(df,markers,"raghavan")
```

```{r, eval=FALSE}
raghavan_t = subset(raghavan_sub, subset = hadaca3 %in% c("classic","basal"))

# agg_expr_raghavan <- AggregateExpression(raghavan_t, features = rob_df$Gene, group.by = "hadaca3")
# zs_raghavan = zs(agg_expr_raghavan$RNA)

df = raghavan_t@assays$RNA$counts
colnames(df) = raghavan_t$hadaca3
  
markers = rob_df

plot_heatmap(df,markers,"raghavan")
```

```{r, eval=FALSE}
# PENG
peng_t = subset(peng_sub, subset = hadaca3 %in% c("classic","basal"))

#subset to 100 basal and 100 classic 
basal = subset(peng_t, hadaca3 == "basal")
classic = subset(peng_t, hadaca3 == "classic")

sample_cl <- sample(Cells(classic), 100)

classic = subset(classic, cells =  sample_cl)

subset = merge(classic,basal)
subset = JoinLayers(subset)

df = subset@assays$RNA$counts
colnames(df) = subset$hadaca3

# purist markers 
markers = data.frame(Compartment = c(rep("Classic",21),rep("Basal",21)),
                    Gene = c(classifier$TSPs[,2],classifier$TSPs[,1]))

plot_heatmap(df,markers,"peng")
```

```{r, eval=FALSE}
# RAGHAVAN
avg_expr_raghavan_log <- log2(avg_expr_raghavan$SCT + 1)
avg_expr_raghavan_scaled <- t(scale(t(avg_expr_raghavan_log)))

zs_raghavan = zs(avg_expr_raghavan_scaled)

# avg_expr_raghavan_scaled[is.na(avg_expr_raghavan_scaled)] <- 0

ann_colors <- list(Classification = c("basal" = "skyblue", "classic" = "salmon"))

# Heatmap(avg_expr_raghavan_scaled,
#         name = "Expression",
#         row_title = "Markers",
#         column_title = "Clusters RAGHAVAN",
#         top_annotation = HeatmapAnnotation(Classification = cluster_classification_raghavan ,
#                                            col = ann_colors))

marker_order = intersect(markers$Gene,rownames(zs_raghavan))

marker_classification <- markers$Compartment[match(marker_order, markers$Gene)]

Heatmap(zs_raghavan,
        name = "Expression",
        row_title = "Markers",
        column_title = "Clusters RAGHAVAN",
        row_order = marker_order, 
        top_annotation = HeatmapAnnotation(Classification = cluster_classification_raghavan ,
                                           col = ann_colors))


# PENG 
avg_expr_peng_log <- log2(avg_expr_peng$SCT + 1)
avg_expr_peng_scaled <- t(scale(t(avg_expr_peng_log)))
# avg_expr_peng_scaled[is.na(avg_expr_peng_scaled)] <- 0

zs_peng = zs(avg_expr_peng_scaled)
# zs_peng[is.na(zs_peng)] = 0

ann_colors <- list(Classification = c("basal" = "skyblue", "classic" = "salmon"))

# Heatmap(avg_expr_peng_scaled,
#         name = "Expression",
#         row_title = "Markers",
#         column_title = "Clusters PENG",
#         top_annotation = HeatmapAnnotation(Classification = cluster_classification_peng ,
#                                            col = ann_colors))
# 
marker_order = intersect(markers$Gene,rownames(zs_peng))
Heatmap(zs_peng,
        name = "Zscore",
        row_title = "Markers",
        column_title = "Clusters PENG",
        row_order = marker_order, 
        top_annotation = HeatmapAnnotation(Classification = cluster_classification_peng ,
                                           col = ann_colors))
```

# Save into list

```{r, results="verbatim"}

sc_rna_ref = list(ref_sc_peng = list(counts=counts_peng, metadata=metadata_peng), ref_sc_baron = list(counts=counts_baron, metadata=metadata_baron), ref_sc_raghavan = list(counts=counts_raghavan, metadata=metadata_raghavan) )

reference_data <- list(ref_bulkRNA=ref_rna, ref_scRNA=sc_rna_ref, ref_met=ref_met)

saveRDS(reference_data, paste0(reference_file))
```







# Session Information

```{r, results="verbatim"}
sessionInfo()
```
