---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
library(MASS) # for fitditrib, compute size parameter for sc source 
# library(pbapply) # for firdistrib and pseuod mthylation sc generation
# library(parallel)

output_dir = "."
ifelse(!dir.exists(file.path(output_dir)), dir.create(file.path(output_dir)), FALSE)


```

```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes1_insilicopseudobulk_pdac.rds","mixes2_insilicopseudobulk_pdac.rds")} 
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth_insilicopseudobulk1_pdac.rds","groundtruth_insilicopseudobulk2_pdac.rds")} 
```

# Objectives

This script generates 3 types of data :

################################

It takes in input src data (script 01.1_ref_XXX.Rmd).

# Read src data 

```{r, results="verbatim", eval=FALSE}

src = readRDS("01_references/source_pdac.rds")

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met
sc_rna_src = src$src_scRNA

rm(src)
gc()

```

# Generate pseudobulk

## Function

```{r }

source("../01-additional/generator.R") # Get generator function from https://github.com/favilaco/deconv_benchmark)
```

```{r }
generate_pseudo_bulk = function(sc_data, phenoData, proportion, Num.mixtures, pool.size, idx_nb , type){ 
    idx=1
    mix = list()

    # Generate the pseudo-bulk
    for (idx in 1:idx_nb) {
        pbmix <- generator_hadaca3(sc_data = sc_data, phenoData = phenoData, proportion=proportion, Num.mixtures = Num.mixtures, pool.size = pool.size,seed = idx, type = type)
        mix[[idx]] <- pbmix # T = tissue
    }

    # Compute the mean metaprofile (equivalent to the T src matrix)
    cellType <- phenoData$cellType
    group = list()
    for(i in unique(cellType)){
        group[[i]] <- which(cellType %in% i)
    }
    C = lapply(group,function(x) Matrix::rowMeans(sc_data[,x]))
    C = round(do.call(cbind.data.frame, C))

    # Save benchmark metadata

    input_k_value = length(unique(phenoData$cellType))
    
    result = c(mix = mix, signature = C)
    return(result)
}
```

## Generate pseudo single cell

### RNA

number of cells in bulk approximate 
```{r, eval=FALSE}

colsum_bulk = colSums(rna_seq_src)
tr_bulk = mean(colsum_bulk) # nb transcripts in bulk
colsum = colSums(as.matrix(sc_rna_src$counts)) 
tr_sc =  mean(colsum) # nb transcripts per cell in sc


tr_bulk/ tr_sc 
```

#### Get size parameter
```{r, eval=FALSE}

CT = colnames(rna_seq_src) #cell types

size_param.list = list()

for (cell_type in CT ) {
idx = which(sc_rna_src$metadata %in% cell_type) # get cells idx for current cell type
count_m_sc = as.matrix(sc_rna_src$counts[,idx])

# Pre-allocate size vector
theta <- numeric(nrow(count_m_sc))

# Loop through each gene and fit glm.nb, with error handling
for (i in seq_len(nrow(count_m_sc))) {
  # checking for all zero counts 
  if (!all(count_m_sc[i, ] == 0)) {
  # Fit the model and handle errors
  model <- tryCatch({
    suppressWarnings(glm.nb(count_m_sc[i, ] ~ 1))
      # glm.nb(count_m_sc[i, ] ~ 1)

  }, error = function(e) {
    
    message("Error raised for row ", i, ": \"", e$message, "\"")
    return(NULL) })  # Return NULL if there's an error
  
  # Store theta if model succeeded, otherwise store NA
  theta[i] <- if (!is.null(model)) model$theta else NA
  } else {
    theta[i] = NA  
  } 
  
 
  # Update progress bar
  percent <- round((i / nrow(count_m_sc)) * 100)
  cat("\rProcessing:", paste(rep("=", percent / 2), collapse = ""), paste0(percent, "%"), sep = " ")
  flush.console()  # Display progress immediately in console
  
}
  #for null model or gene with all counts = 0, randomly pick a size the the pool of generated size 
  
  theta[is.na(theta)] = sample(x = theta[!is.na(theta)], size = length(theta[is.na(theta)]))
 
  # Store size parameters for the current cell type
  size_param.list[[cell_type]] <- theta
  
  # Print completion message for the cell type
  cat("\nProcessing complete for cell type:", cell_type, "\n")
}


size_param_df = as.data.frame(size_param.list)
rownames(size_param_df) = rownames(sc_rna_src$counts) # gene names
saveRDS(size_param_df,"01.2_size_praram_sc_src_glmnb.rds")
```

#### Generate sc RNA 

```{r, eval=FALSE}
# size_param_df = readRDS("01.2_size_praram_sc_src_glmnb.rds")

common_genes = intersect(rownames(sc_rna_src$counts) , rownames(rna_seq_src) ) #get the common gene bewteen the sincgle cell and bulk rna 

# generate new pseudo single cell data using parameters from peng and rna_seq_src 

cell_types = colnames(size_param_df)
ncells = 100 #size of simulation for each cell type 
pseudo_sc_rna = list()
metadata_cell_type_psc_rna = c()

#for each gene get a random count value from a 
# - normal distribution with mean from rna_seq_src and sd from peng
# - negative binomial distribution

set.seed(123)

nb_bulk_cells = 5000 # arbitrary value based on the previous chunk 

for (cell_type in cell_types) { 
  # Initialize the matrix for this cell type
  matrix <- matrix(nrow = length(common_genes), ncol = ncells)

  print(cell_type)

  # Inner loop over common_genes with progress tracking
  for (i in seq_along(common_genes)) {

    gene <- common_genes[i]

    # Calculate and display loading bar
    # percent <- round((i / length(common_genes[1:100])) * 100)
    percent <- round((i / length(common_genes)) * 100)

    cat("\rProgress:", paste(rep("=", percent), collapse = ""), paste0(percent, "%"), sep = " ")
    flush.console() # Flush console to immediately display progress

    # Core calculations
    counts <- list()
    mean <- rna_seq_src[gene, cell_type] / nb_bulk_cells
    size <- size_param_df[gene, cell_type]
    #size = mean^2/(sd^2-mean)

    # Generate counts using the negative binomial distribution
    counts <- rnbinom(n = ncells, size = size, mu = mean)

    # Store the result in the matrix at the correct index
    matrix[i, ] <- counts
  }
  
  cat("\n") # Print newline after loading bar completes
  
  # Set row names for the matrix
  rownames(matrix) <- common_genes

  # Update metadata and store the matrix
  metadata_cell_type_psc_rna <- c(metadata_cell_type_psc_rna, rep(cell_type, ncells))
  pseudo_sc_rna[[cell_type]] <- matrix
  print(paste(cell_type, "DONE"))
}


pseudo_sc_rna_count_matrix = do.call(cbind,pseudo_sc_rna)
colnames(pseudo_sc_rna_count_matrix) = make.names(rep("cell",dim(pseudo_sc_rna_count_matrix)[2]),unique = TRUE) #create random cell ID 

saveRDS(pseudo_sc_rna_count_matrix,"pseudo_sc_rna_count_matrix.rds")
saveRDS(metadata_cell_type_psc_rna,"metadata_cell_type_psc_rna.rds")
```

### Met

Custom propability law: 

Let p0, p0.5, and p1 be the probabilities of generating 0, 0.5, and 1, respectively:

    p0 is the probability of selecting 0.
    p0.5 is the probability of selecting 0.5.
    p1 is the probability of selecting 1.

We have the following constraints:

    Sum of probabilities:
    p0+p0.5+p1=1
    Mean of the distribution:
    0*p0+0.5*p0.5+1*p1=μ
Solution:

    From the mean constraint:
    0.5p0.5+p1=μ
    
    p0.5=2μ−2p1


    Substitute p0.5 in the sum of probabilities:
    p0+2μ−2p1+p1=1

    p0+2μ−p1=1

    p0=1−2μ+p1


Choosing p1 to get valid probabilities:

    To ensure p0, p0.5, and p1 are all non-negative and sum to 1, p1 should satisfy:
    0≤1−2μ+p1≤1

    0≤2μ−2p1≤1


This gives:
max⁡(0,2μ−1)≤p1≤min⁡(2μ,1)


If p1 is too close to the upper bound of 1 then we risk having p0.5 negative.

 To avoid this, we need to impose an additional constraint to ensure p0.5 stays non-negative.
Updated approach:

    The valid range for p1 is not only determined by μ, but also by ensuring that p0.5≥0, which requires:
    p1≤μ

    The valid range for p1 thus becomes:
    max⁡(0,2μ−1)≤p1≤min⁡(μ,2μ)

```{r, eval=FALSE}


probe_400k = unlist(as.vector(read.csv("~/projects/datashare/genref_hadaca3/cpg_probes_400k", header = F))) # list of probes from GSE82234 GEO dataset

# down size dna met datset 

dna_met_src = dna_met_src[rownames(dna_met_src) %in% probe_400k,]

# generate new pseudo single cell data using parameters from peng and rna_seq_src 

cell_types = colnames(dna_met_src)
ncells = 100 #size for each cell type 
probes = rownames(dna_met_src)

pseudo_sc_met = list()
metadata_cell_type_psc_met = c()

for (cell_type in cell_types) { 
  # matrix <- matrix(nrow = 2000, ncol = ncells)
  matrix = matrix(nrow = dim(dna_met_src)[1] ,ncol = ncells)
  print(cell_type)

  # Inner loop over probes with progress tracking
  for (i in seq_along(probes))
    {
    probe <- probes[i]
    
    # Calculate and display loading bar
    percent <- round((i / length(probes)) * 100)
    cat("\rProgress:", paste(rep("=", percent), collapse = ""), paste0(percent, "%"), sep = " ")
    flush.console() # Flush console to immediately display progress
    
    # Core calculations
    met <- list()
    mu <- dna_met_src[probe, cell_type] 
    
    # Define valid range for p1
    min_p1 <- max(0, 2 * mu - 1)
    max_p1 <- min(mu, 2 * mu)
    
    # Choose p1 within the valid range
    p1 <- runif(n = 1, min = min_p1, max = max_p1)    
    # Calculate p0 and p0.5 based on p1 and mu
    p0 <- 1 - 2 * mu + p1
    p0.5 <- 2 * mu - 2 * p1

    # Define a small tolerance for floating-point comparison
    tolerance <- 1e-9
    
    # Check that the probabilities are valid with a tolerance
    if (p0 < 0 || p0.5 < 0 || p1 < 0 || abs(p0 + p0.5 + p1 - 1) > tolerance) {
      stop("Invalid probabilities")
    }
    
    # Sample based on probabilities
    met <- sample(c(0, 0.5, 1), size = ncells, prob = c(p0, p0.5, p1), replace = TRUE)

    # Store result in matrix at correct index
    matrix[i, ] <- met
  }
  
  cat("\n") # Print newline after loading bar completes
  
  # Set rownames for the matrix
  rownames(matrix) = probes


  # Update metadata and store matrix
  metadata_cell_type_psc_met <- c(metadata_cell_type_psc_met, rep(cell_type, ncells))
  pseudo_sc_met[[cell_type]] <- matrix
  print(paste(cell_type, "DONE"))
}


pseudo_sc_met_count_matrix = do.call(cbind,pseudo_sc_met)
colnames(pseudo_sc_met_count_matrix) = make.names(rep("cell",dim(pseudo_sc_met_count_matrix)[2]),unique = TRUE) #create random cell ID 

# saveRDS(pseudo_sc_met_count_matrix,"pseudo_sc_met_count_matrix.rds")
# saveRDS(metadata_cell_type_psc_met,"metadata_cell_type_psc_met.rds")
```

## Proportion matrix generation
 
```{r, results="verbatim"}
# set.seed(2)

simu_proportion_pancreas = function(n, alpha = c(1.5, 4.5, 1, 3)){
  tmp_mix_Lvar = t(gtools::rdirichlet(n = n, alpha = alpha))
  tumor_rep = cbind(t(gtools::rdirichlet(n = n/3, alpha = c(0.5, 0.5))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.75, 0.25))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.9, 0.1))))
  mix_Lvar = matrix(nrow = 5, ncol = n)
  mix_Lvar [1:3, ] = tmp_mix_Lvar[1:3, ]
  mix_Lvar[4, ] = tumor_rep[1, ]*tmp_mix_Lvar[4, ]
  mix_Lvar[5, ] = tumor_rep[2, ]*tmp_mix_Lvar[4, ]

  return(mix_Lvar)
}

nb_samples = 60

proportion = simu_proportion_pancreas(n = nb_samples)
rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") 
```

## Ground truth - split {-}

```{r prop_mat_saving, results="verbatim"}
set.seed(2024)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```

## Generate pseudo bulk from single Cell 

### Directly with peng data 

```{r, eval=FALSE}
cur = sc_rna_src
phenoData = data.frame(cellType = as.vector(cur$metadata),
                       cellID = rownames(cur$metadata),
                       sampleID = rep("sample",dim(cur$metadata)[2])) #create artificial sample
sc_data = cur$counts
type = "rna"


gen_peng = generate_pseudo_bulk(sc_data = sc_data,
					phenoData = phenoData,
					Num.mixtures = Num.mixtures,
					pool.size = pool.size,
					idx_nb = idx_nb,
					type = type)
```


```{r}
input_dir = "~/projects/datashare/pseudobulk_mix_hadaca3/"

pseudo_sc_rna_count_matrix = readRDS(paste0(input_dir, "pseudo_sc_rna_count_matrix.rds"))
pseudo_sc_met_count_matrix = readRDS(paste0(input_dir, "pseudo_sc_met_count_matrix.rds"))
metadata_cell_type_psc_rna = readRDS(paste0(input_dir, "metadata_cell_type_psc_rna.rds"))
metadata_cell_type_psc_met = readRDS(paste0(input_dir, "metadata_cell_type_psc_met.rds"))

```



###  With generated pseudo single cell RNA

```{r}
phenoData = data.frame(cellType = metadata_cell_type_psc_rna,
                       cellID = colnames(pseudo_sc_rna_count_matrix),
                       sampleID = rep("sample",dim(pseudo_sc_rna_count_matrix)[2])) #create artificial sample

sc_data = pseudo_sc_rna_count_matrix
type = "rna"
dir_name = "data_pseudo_bulk_psc_rna" #psc = pseudo single cell

# parameters for pseudobulk generation function 
# Num.mixtures = 50 #nb of sample generated
pool.size = 100 #number of cells used from single cell data set to generate bulk counts
idx_nb = 1 #nb simulation, (nb of bulk matrices)

gen_rna = generate_pseudo_bulk(sc_data = sc_data,
					phenoData = phenoData,
					proportion = proportion,
					# Num.mixtures = Num.mixtures,
					Num.mixtures = nb_samples, # from proportion matrix generation
					pool.size = pool.size,
					idx_nb = idx_nb,
					type = type)

```

### With generated pseudo single cell met


```{r}
phenoData = data.frame(cellType = metadata_cell_type_psc_met,
                       cellID = colnames(pseudo_sc_met_count_matrix),
                       sampleID = rep("sample",dim(pseudo_sc_met_count_matrix)[2])) #create artificial sample

sc_data = pseudo_sc_met_count_matrix
type = "met"
dir_name = "data_pseudo_bulk_psc_met" #psc = pseudo single cell



# parameters for pseudobulk generation function 
# Num.mixtures = 50 #nb of sample generated
pool.size = 100 #number of cells used from single cell data set to generate bulk counts
idx_nb = 1 #nb simulation, (nb of bulk matrices)

gen_met = generate_pseudo_bulk(sc_data = sc_data,
					phenoData = phenoData,
					proportion = proportion,
					# Num.mixtures = Num.mixtures,
					Num.mixtures = nb_samples, # from proportion matrix generation
					pool.size = pool.size,
					idx_nb = idx_nb,
					type = type)

```

## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=gen_rna$mix[,idx_phaseUsers],
                              mix_met=gen_met$mix[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=gen_rna$mix[,-idx_phaseUsers],
                              mix_met=gen_met$mix[,-idx_phaseUsers])

```


# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```
