---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
```


```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes1_insilicodirichletnorm_pdac.rds","mixes2_insilicodirichletnorm_pdac.rds")}
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth1_insilicodirichletnorm_pdac.rds","groundtruth2_insilicodirichletnorm_pdac.rds")}
```

# Objectives

This script generates :

- mixes_data.rds (a list with `mix_rna` and `mix_met` matrices), 

- ground_truth.rds

It takes as input source data (script 01.1_src_XXX.Rmd).


# Read source data 

```{r, results="verbatim"}
src = readRDS(paste0("../",source_file))

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met

rm(src)
gc()
```

# Proportion matrix simulation

An example of generative model for 5 cell types, in pancreatic adenocarcinoma.

General distribution :

    normal 15%
    fibro 45%,
    immune 10%,
    cancer 30% (divided between basal and classical subtypes)
    
Subdivision of cancer types among cancer cells.

-> in 1/3 of the samples : 50% type basal, 50% type classical

-> in 1/3 of the samples : 25% type basal, 75% type classical

-> in 1/3 of the samples : 10% type basal, 90% type classical


## Functions 

```{r, results="verbatim"}
simu_proportion_pancreas = function(n, alpha = c(1.5, 4.5, 1, 3)){
  tmp_mix_Lvar = t(gtools::rdirichlet(n = n, alpha = alpha))
  tumor_rep = cbind(t(gtools::rdirichlet(n = n/3, alpha = c(0.5, 0.5))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.75, 0.25))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.9, 0.1))))
  mix_Lvar = matrix(nrow = 5, ncol = n)
  mix_Lvar [1:3, ] = tmp_mix_Lvar[1:3, ]
  mix_Lvar[4, ] = tumor_rep[1, ]*tmp_mix_Lvar[4, ]
  mix_Lvar[5, ] = tumor_rep[2, ]*tmp_mix_Lvar[4, ]

  return(mix_Lvar)
}
```

## Proportion matrix generation

Below is the code to generate a list of 'length.out' simulations with 'n' samples

```{r, results="verbatim", eval=FALSE}
set.seed(10)
ground_truth_list <- lapply(
  X = seq_len(length.out = 5) # length of X = number of simulation instances
  , FUN = function(i) {
    proportion = simu_proportion_pancreas(n = 30) # n -> number of samples
    rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
    return(proportion)
  }
)
```

Below is the code to generate only 1 simulation 

```{r, results="verbatim"}
set.seed(10)

nb_samples = 60

proportion = simu_proportion_pancreas(n = nb_samples) # n -> number of samples
rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
```

## Ground truth - split {-}

```{r}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```


# Simulation

2 options : 

* convolution of source data using the proportion matrix

* pseudo-bulk using the proportion matrix
Generate pseudo-single cells (generate 1000 pseudo-single-cells by pure cell type by inferring variance parameters from scRNA-seq data)
  
## Simulation of DNAm and RNA mixes from convolution (Mix_met, Mix_RNA)

The following function adds noise to the simulation to make the deconvolution harder and the simulation more realistic.

```{r, results="verbatim"}
# function to add noise on D matrix
add_noise = function(data, mean = 0, sd = 0.05, val_min = 0, val_max = 1) {  # val min et val max make sure the value stays biologicaly coherent ([0:1] for met and [0:...] for rna)
  noise = matrix(rnorm(prod(dim(data)), mean = mean, sd = sd), nrow = nrow(data)) 
  datam = data + noise
  datam[datam < val_min] = data[datam < val_min]
  datam[datam > val_max] = data[datam > val_max]
  return(datam)
}
```

### Simulation Mix_RNA

Below is the code to generate a list of 'length.out' simulations with 'n' samples

```{r, results="verbatim", eval=FALSE}
set.seed(20)

Mix_RNA <- lapply(
  X = seq_len(length.out = 5) # number of simulations
  , FUN = function(i) {
RNA_brut = as.matrix(rna_seq_src) %*%  ground_truth_list[[i]] #convolution of source matrix with proportion matrix
RNA = add_noise(RNA_brut, sd = 20, val_min = 0, val_max = 10000)  # add noise to harden deconvolution, to mimic technical noise/biological noise, test sd value fit for rna bulk 
  }
)

head(public_data_rna[[1]])
```

Below is the code to generate only 1 simulation 

```{r, results="verbatim"}
set.seed(20)

Mix_RNA_brut =  as.matrix(rna_seq_src) %*% proportion #convolution of source matrix with proportion matrix
Mix_RNA = add_noise(Mix_RNA_brut, sd = 20, val_min = 0, val_max = 10000)  # add noise to harden deconvolution, and to mimic technical noise

head(Mix_RNA)
```

### Simulation Mix_met

Below is the code to generate a list of 'length.out' simulations with 'n' samples

```{r, results="verbatim", eval=FALSE}
set.seed(20)

Mix_met <- lapply(
  X = seq_len(length.out = 5)
  , FUN = function(i) {
     met_brut =  as.matrix(dna_met_src) %*% ground_truth_list[[i]] #convolution of source matrix with proportion matrix
    met = add_noise(met_brut, sd = 0.2, val_min = 0, val_max = 1)  # add noise to harden deconvolution, to mimic technical noise/biological noise
  }
)

head(public_data_met[[1]])
```

Below is the code to generate only 1 simulation 

```{r, results="verbatim"}
set.seed(20)

Mix_met_brut = as.matrix(dna_met_src) %*% proportion #convolution of source matrix with proportion matrix
Mix_met = add_noise(Mix_met_brut, sd = 0.2, val_min = 0, val_max = 1)  # add noise to harden deconvolution, to mimic technical noise

head(Mix_met)
```


## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=Mix_RNA[,idx_phaseUsers],
                              mix_met=Mix_met[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=Mix_RNA[,-idx_phaseUsers],
                             mix_met=Mix_met[,-idx_phaseUsers])
```

# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```
