---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe, Hugo Barbot and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)
# library(Seurat)
```


```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes_insilicodirichletEMFA1_pdac.rds","mixes_insilicodirichletEMFA2_pdac.rds")} 
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth_insilicodirichletEMFA1_pdac.rds","groundtruth_insilicodirichletEMFA2_pdac.rds")} 
```

# Objectives

This script generates :

- mixes_data.rds (a list with `mix_rna` and `mix_met` matrices), 

- ground_truth.rds

It takes in input source data (script 01.1_src_XXX.Rmd).

# Read source data 

```{r, results="verbatim"}
src = readRDS(paste0("../",source_file))

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met
rm(src)
gc()

```

# Proportion matrix simulation

An example of generative model for 5 cell types, in pancreatic adenocarcinoma.

General distribution :

    normal 15%
    fibro 45%,
    immune 10%,
    cancer 30% (divided between basal and classic)
    
Subdivision of cancer types among cancer cells.

-> in 1/3 samples : 50% type basal, 50% type classic

-> in 1/3 samples : 25% type basal, 75% type classic

-> in 1/3 samples : 10% type basal, 90% type classic




## Proportion matrix generation

Below code to generate only 1 simulation 

```{r prop_mat_generation, results="verbatim"}
# set.seed(1)

nb_samples = 60



# alpha_prop1 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45, 0.1, 0.5*0.3, 0.5*0.3), # 0.15, 0.15
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop1 = c(5.162500, 15.487500, 3.441667, 5.162500, 5.162500)

# alpha_prop2 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.75*0.3, 0.25*0.3), # 0.225, 0.075
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop2 = c(5.162500, 15.487500, 3.441667, 7.743750, 2.581250)

# alpha_prop3 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.9*0.3, 0.1*0.3), # 0.27, 0.03
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop3 = c(5.162500, 15.487500, 3.441667, 9.292500, 1.032500)


n = floor(nb_samples/3)
proportion = t(rbind(
  gtools::rdirichlet(n = n + (nb_samples - 3*n), alpha = alpha_prop1),
  gtools::rdirichlet(n = n, alpha = alpha_prop2),
  gtools::rdirichlet(n = n, alpha = alpha_prop3)
))

rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!

rm(alpha_prop1, alpha_prop2, alpha_prop3, n)
```

```{r, results="verbatim", eval=FALSE}
set.seed(10)

simu_proportion_pancreas = function(n, alpha = c(1.5, 4.5, 1, 3)){
  tmp_mix_Lvar = t(gtools::rdirichlet(n = n, alpha = alpha))
  tumor_rep = cbind(t(gtools::rdirichlet(n = n/3, alpha = c(0.5, 0.5))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.75, 0.25))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.9, 0.1))))
  mix_Lvar = matrix(nrow = 5, ncol = n)
  mix_Lvar [1:3, ] = tmp_mix_Lvar[1:3, ]
  mix_Lvar[4, ] = tumor_rep[1, ]*tmp_mix_Lvar[4, ]
  mix_Lvar[5, ] = tumor_rep[2, ]*tmp_mix_Lvar[4, ]

  return(mix_Lvar)
}

nb_samples = 60

proportion = simu_proportion_pancreas(n = nb_samples) # n -> number of samples
rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
```

## Ground truth - split {-}

```{r prop_mat_saving, results="verbatim"}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```

  
# Simulation of DNAm and RNA mixes from convolution Mix_met Mix_RNA

The following function adds noise to the simulation to make is harder to deconvolve, more life like.

In order to make the deconvolution more difficult, here the added noise is randomly chosen from distribution with different parameters for different randomly selected subset of genes/met probes. 

Laplace or chi2 law can be used. Issue of chi2 is that it only add positive values. 


## Simulation Mix_RNA

Below code to generate only 1 simulation 

```{r RNA_data_driven_noise, results="verbatim"}
emfa_sc_epsilon_in_vitro_rna = readRDS(file = "../01-additional/simu_data_driven/emfa_sc_epsilon_in_vitro_rna_21000.rds")
q = 20 # number of factor use in emfa
mean_epsilon_in_vitro_rna = readRDS(file = "../01-additional/simu_data_driven/mean_epsilon_in_vitro_rna_21000.rds")
sd_epsilon_in_vitro_rna = readRDS(file = "../01-additional/simu_data_driven/sd_epsilon_in_vitro_rna_21000.rds")


E = tcrossprod(rnorm(length(sd_epsilon_in_vitro_rna), 0, emfa_sc_epsilon_in_vitro_rna$Psi), rep(1,nb_samples)) # mean_epsilon_in_vitro_rna

Z = MASS::mvrnorm(nb_samples, mu = rep(0, q), Sigma = diag(1, ncol = q, nrow = q))

noise_rna_aggr = ( E + tcrossprod(emfa_sc_epsilon_in_vitro_rna$B, Z) ) * sd_epsilon_in_vitro_rna

# This methodology simulate scaled and center epsilon errors, so I add a "bias" (in the matrix E) and I multiply be the sd I want for each gene
```

```{r adding_Mix_RNA_and_noise, results="verbatim"}
Mix_RNA = as.matrix(rna_seq_src) %*% proportion #convolution of source 

Mix_RNA = Mix_RNA + noise_rna_aggr

which_gene_neg = apply(Mix_RNA, 1, \(gene_i){any(gene_i<0)})
which_gene_neg = names(which_gene_neg[which(which_gene_neg)])

for (gene_i in which_gene_neg) {
  # ddl_gene_i = sd_epsilon_in_vitro_rna[gene_i]
  Mix_RNA[gene_i,] = Mix_RNA[gene_i,] - min(Mix_RNA[gene_i,]) + runif(nb_samples, min = 0, max = 1e-3)
}

# tmp = Mix_RNA - as.matrix(rna_seq_src) %*% proportion # noise_rna_aggr
# tmptmp = cor(t(tmp))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of residual correlations for simulation by emfa (rna)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

rm(emfa_sc_epsilon_in_vitro_rna, sd_epsilon_in_vitro_rna, E, Z, noise_rna_aggr, which_gene_neg)
gc()
```


## Simulation Mix_met

```{r aggregated_MET_data_driven_noise, results="verbatim"}
emfa_sc_epsilon_in_vitro_met = readRDS(file = "../01-additional/simu_data_driven/emfa_sc_epsilon_in_vitro_met.rds")
q = 20 # number of factor use in emfa
mean_epsilon_in_vitro_met = readRDS(file = "../01-additional/simu_data_driven/mean_epsilon_in_vitro_met.rds")
sd_epsilon_in_vitro_met = readRDS(file = "../01-additional/simu_data_driven/sd_epsilon_in_vitro_met.rds")


E = tcrossprod(rnorm(length(sd_epsilon_in_vitro_met), 0, emfa_sc_epsilon_in_vitro_met$Psi), rep(1,nb_samples)) # mean_epsilon_in_vitro_met

Z = MASS::mvrnorm(nb_samples, mu = rep(0, q), Sigma = diag(1, ncol = q, nrow = q))

noise_met_aggr = ( E + tcrossprod(emfa_sc_epsilon_in_vitro_met$B, Z) ) * sd_epsilon_in_vitro_met

rm(emfa_sc_epsilon_in_vitro_met, sd_epsilon_in_vitro_met, E, Z)
gc()
```

Transformation of probes.features to know the region and the gene of where a probe is attached, but we don't have access to the true proportion here for pdac data
```{r aggregate_to_820k, eval=FALSE}
probes_feature = readRDS("../01-additional/probes_features.rds")

probes_feature = probes_feature %>% dplyr::select(CHR, gene, feature)  %>% 
  mutate(
    feature_promoter = case_when(
      feature %in% c("TSS200", "TSS1500") ~ "Promoter",
      # feature == "Body" ~ feature,
      feature == "IGR" ~ feature, # if we are between gene, we keep this information
      #(even if we merge after so don't have this information anymore)
      TRUE ~ "Other" # only the promoter region have an interest here
    ),
    feature_promoter = factor(feature_promoter, levels = c("Promoter", "IGR", "Other"))
  ) %>% 
  rownames_to_column(var = "CpG_site")
# probes_feature %>% group_by(gene, feature_promoter) %>% summarise(n = n())


# En utilisant les données de methylation pure, puis en calculant les erreurs (matrice de 820000 par 30)

sd_per_gene_from_CpG = left_join(
  epsilon_in_vitro_met %>% as.data.frame() %>% rownames_to_column(var = "CpG_site"),
  probes_feature,
  by = "CpG_site"
) %>% 
  group_by(gene, feature_promoter) %>% 
  summarise(
    n = n(),
    
    sd_01 = sd(COMT01TFZP), sd_02 = sd(COMT02TFZP), sd_03 = sd(COMT03TFZP),
    sd_04 = sd(COMT04TFZP), sd_05 = sd(COMT05TFZP), sd_06 = sd(COMT06TFZP),
    sd_07 = sd(COMT07TFZP), sd_08 = sd(COMT08TFZP), sd_09 = sd(COMT09TFZP),
    sd_10 = sd(COMT10TFZP), sd_11 = sd(COMT11TFZP), sd_12 = sd(COMT12TFZP),
    sd_13 = sd(COMT13TFZP), sd_14 = sd(COMT14TFZP), sd_15 = sd(COMT15TFZP),
    sd_16 = sd(COMT16TFZP), sd_17 = sd(COMT17TFZP), sd_18 = sd(COMT18TFZP),
    sd_19 = sd(COMT19TFZP), sd_20 = sd(COMT20TFZP), sd_21 = sd(COMT21TFZP),
    sd_22 = sd(COMT22TFZP), sd_23 = sd(COMT23TFZP), sd_24 = sd(COMT24TFZP),
    sd_25 = sd(COMT25TFZP), sd_26 = sd(COMT26TFZP), sd_27 = sd(COMT27TFZP),
    sd_28 = sd(COMT28TFZP), sd_29 = sd(COMT29TFZP), sd_30 = sd(COMT30TFZP)
  )


sd_per_gene_from_CpG = sd_per_gene_from_CpG %>% 
  rowwise() %>% 
  mutate(mean_sd = sd(c_across(sd_01:sd_30))) %>% 
  dplyr::select(gene, feature_promoter, n, mean_sd)


list_arg_aggr_to_820k = left_join(
  left_join(
    epsilon_in_vitro_met %>% as.data.frame() %>% rownames_to_column(var = "CpG_site") %>% dplyr::select(CpG_site),
    probes_feature %>% dplyr::select(CpG_site, gene, feature_promoter),
    by = "CpG_site"
  ),
  sd_per_gene_from_CpG,
  by = c("gene", "feature_promoter")
) 

# write_rds(list_arg_aggr_to_820k, file = "list_arg_aggr_to_820k.rds")
```


```{r duplicate_and_randomise_noise_to_820k_probes, results="verbatim"}
list_arg_aggr_to_820k = readRDS(file = "../01-additional/simu_data_driven/list_arg_aggr_to_820k.rds")


# set.seed(2)

Mix_met = as.matrix(dna_met_src) %*% proportion #convolution of source matrix with proportion matrix

## separated probes in subsets of those where their gene is in the aggregated and paired data and others where i use chi2 noise from insilisodirichletchi2correl 


list_known_sites = dplyr::left_join(
  data.frame(gene = rownames(noise_met_aggr)),
  list_arg_aggr_to_820k,
  by = "gene"
)
# recover only probes in Mix_Met (as match)
list_known_sites = list_known_sites[which(list_known_sites$CpG_site %in% rownames(Mix_met)),]
# match(list_known_sites$CpG_site, rownames(Mix_met))


list_unknown_sites = list_arg_aggr_to_820k[!(list_arg_aggr_to_820k$CpG_site %in% list_known_sites$CpG_site),]
# recover only probes in Mix_Met (as match)
list_unknown_sites = list_unknown_sites[which(list_unknown_sites$CpG_site %in% rownames(Mix_met)),]


## Add noise different noise to first subset of gene, those who I have generate "aggregated" noise


noise_met_known_sites = dplyr::left_join(
  list_known_sites[, c("CpG_site", "gene")],
  tibble::rownames_to_column(as.data.frame(noise_met_aggr), var = "gene") ,
  by = "gene"
)
noise_met_known_sites = tibble::column_to_rownames(noise_met_known_sites, var = "CpG_site")[,2:(nb_samples+1)]# -c("gene")

probes = list_known_sites$CpG_site[order(list_known_sites$CpG_site)]
Mix_met_pure_noise_known_sites = Mix_met[probes,] + noise_met_known_sites[probes,]


list_probe_mean = as.vector(t(as.matrix( # take the transposed matrix because i want line by line (but we can repeat sd vector 30 times rather than repeat each entries 30 times)
  Mix_met_pure_noise_known_sites 
)))
list_probe_mean[list_probe_mean <= 0] = runif(sum(list_probe_mean <= 0), min = 1e-06, max = 1e-04)
list_probe_mean[list_probe_mean >= 1] = runif(sum(list_probe_mean >= 1), min = 1-1e-04, max =  1-1e-06)

list_probe_sd = rep(list_known_sites$mean_sd, each = nb_samples)


probe_param_tmp = list_probe_mean * (1 - list_probe_mean) / list_probe_sd^2
probe_a = list_probe_mean * (probe_param_tmp - 1)
probe_b = (1 - list_probe_mean) * (probe_param_tmp - 1)


Mix_met_known_sites = rbeta(length(list_probe_mean), shape1 = probe_a, shape2 = probe_b)

# If some na are generated, we take the pure noise version
Mix_met_known_sites[is.na(Mix_met_known_sites)] = list_probe_mean[is.na(Mix_met_known_sites)] # Mix_met_pure_noise_known_sites

# Order of probes is conserved with byrow = TRUE
Mix_met_known_sites = matrix(Mix_met_known_sites, ncol = nb_samples, byrow = TRUE)
rownames(Mix_met_known_sites) = probes



## Add noise different noise to second subset of probe,those in an intergenic region

# # Strat 1: calculate a "pure" and mean noise to serve as an "aggregated" noise from the in vitro dataset
# met_measure = readRDS("projects/datashare/data_cometh_lot1/test_data_met_new.rds")[,1:30]
# ground_truth = cbind(readRDS("groundtruth_invitro1_pdac.rds"),
#                      readRDS("groundtruth_invitro2_pdac.rds"))
# ground_truth = ground_truth[,order(colnames(ground_truth))]
# 
# unknown_probes = list_unknown_sites$CpG_site
# pure_noise_unknown_sites = met_measure[unknown_probes,] - dna_met_src[unknown_probes,] %*% ground_truth
# rm(met_measure)
# rm(ground_truth)
# 
# Mix_met_pure_noise_unknown_sites = Mix_met[unknown_probes,] + pure_noise_unknown_sites[unknown_probes,]
# 
# list_probe_IGR_mean = as.vector(t(as.matrix( # take the transposed matrix because i want line by line (but we can repeat sd vector 30 times rather than repeat each entries 30 times)
#   Mix_met_pure_noise_unknown_sites 
# )))
# list_probe_IGR_mean[list_probe_IGR_mean <= 0] = runif(sum(list_probe_IGR_mean <= 0), min = 1e-06, max = 1e-04)
# list_probe_IGR_mean[list_probe_IGR_mean >= 1] = runif(sum(list_probe_IGR_mean >= 1), min = 1-1e-04, max =  1-1e-06)
# probe_sd_IGR = unique(list_arg_aggr_to_820k[list_arg_aggr_to_820k$feature_promoter == "IGR", "mean_sd"]) # normalement une seule valeur
# 
# probe_IGR_param_tmp = list_probe_IGR_mean * (1 - list_probe_IGR_mean) / probe_sd_IGR^2
# probe_IGR_a = list_probe_IGR_mean * (probe_IGR_param_tmp - 1)
# probe_IGR_b = (1 - list_probe_IGR_mean) * (probe_IGR_param_tmp - 1)
# 
# 
# Mix_met_unknown_sites = rbeta(length(list_probe_IGR_mean), shape1 = probe_IGR_a, shape2 = probe_IGR_b)
# 
# # If some na are generated, we take the pure noise version
# Mix_met_unknown_sites[is.na(Mix_met_unknown_sites)] = list_probe_IGR_mean[is.na(Mix_met_unknown_sites)] # Mix_met_pure_noise_known_sites
# 
# # Order of probes is conserved with byrow = TRUE
# Mix_met_unknown_sites = matrix(Mix_met_unknown_sites, ncol = nb_samples, byrow = TRUE)
# rownames(Mix_met_unknown_sites) = unknown_probes

# Strat 2: we sample Mix_met_pure_noise_known_sites for unknown sites and generate with the IGR standard deviation. I think it's not a wrong hypothesis to consider IGR probes "attached" to a gene

probes_sample_from_known_sites = sample(1:length(probes), size = nrow(list_unknown_sites), replace = TRUE)
Mix_met_pure_noise_unknown_sites = Mix_met_pure_noise_known_sites[probes_sample_from_known_sites,]


list_probe_IGR_mean = as.vector(t(as.matrix( # take the transposed matrix because i want line by line (but we can repeat sd vector 30 times rather than repeat each entries 30 times)
  Mix_met_pure_noise_unknown_sites 
)))
list_probe_IGR_mean[list_probe_IGR_mean <= 0] = runif(sum(list_probe_IGR_mean <= 0), min = 1e-06, max = 1e-04)
list_probe_IGR_mean[list_probe_IGR_mean >= 1] = runif(sum(list_probe_IGR_mean >= 1), min = 1-1e-04, max =  1-1e-06)

probe_IGR_sd = unique(list_arg_aggr_to_820k[list_arg_aggr_to_820k$feature_promoter == "IGR", "mean_sd"]) # normalement une seule valeur


probe_IGR_param_tmp = list_probe_IGR_mean * (1 - list_probe_IGR_mean) / probe_IGR_sd^2
probe_IGR_a = list_probe_IGR_mean * (probe_IGR_param_tmp - 1)
probe_IGR_b = (1 - list_probe_IGR_mean) * (probe_IGR_param_tmp - 1)


Mix_met_unknown_sites = rbeta(length(list_probe_IGR_mean), shape1 = probe_IGR_a, shape2 = probe_IGR_b)

# If some na are generated, we take the pure noise version
Mix_met_unknown_sites[is.na(Mix_met_unknown_sites)] = list_probe_IGR_mean[is.na(Mix_met_unknown_sites)] # Mix_met_pure_noise_known_sites

# Order of probes is conserved with byrow = TRUE
Mix_met_unknown_sites = matrix(Mix_met_unknown_sites, ncol = nb_samples, byrow = TRUE)
rownames(Mix_met_unknown_sites) = list_unknown_sites$CpG_site[order(list_unknown_sites$CpG_site)]




# replacing_Mix_met

known_probes = list_known_sites$CpG_site[order(list_known_sites$CpG_site)]
Mix_met[known_probes,] = Mix_met_known_sites[known_probes,]

unknown_probes = list_unknown_sites$CpG_site
# Mix_met = add_noise_chi2(unknown_probes, Mix_met, df = 0.1, val_min = 0, val_max = 1) # ne me convient pas
Mix_met[unknown_probes,] = Mix_met_unknown_sites[unknown_probes,]


# Normally not usefull because we use Beta randomness
Mix_met[Mix_met<0] = runif(sum(Mix_met<0), min = 0, max = 1e-03)
Mix_met[Mix_met>1] = runif(sum(Mix_met>1), min = 1-1e-03, max = 1)

# tmp_sample = sample(1:nrow(Mix_met), 20000, replace = FALSE)
# tmp = Mix_met[tmp_sample,] - as.matrix(dna_met_src[tmp_sample,]) %*% proportion
# tmptmp = cor(t(tmp))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of residual correlations for simulation by emfa (met)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

rm(list_arg_aggr_to_820k, list_known_sites, list_unknown_sites, noise_met_known_sites, Mix_met_pure_noise_known_sites, list_probe_mean, list_probe_sd, probe_param_tmp, probe_a, probe_b, Mix_met_known_sites, probes_sample_from_known_sites, Mix_met_pure_noise_unknown_sites, list_probe_IGR_mean, probe_IGR_sd, probe_IGR_param_tmp, probe_IGR_a, probe_IGR_b, Mix_met_unknown_sites, xx)
```

## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=Mix_RNA[,idx_phaseUsers],
                              mix_met=Mix_met[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=Mix_RNA[,-idx_phaseUsers],
                             mix_met=Mix_met[,-idx_phaseUsers])
```


# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```

