---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
```


```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes1_insilicodirichletchi2correl_pdac.rds","mixes2_insilicodirichletchi2correl_pdac.rds")} 
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth1_insilicodirichletchi2correl_pdac.rds","groundtruth2_insilicodirichletchi2correl_pdac.rds")} 
```

# Objectives

This script generates :

- mixes_data.rds (a list with `mix_rna` and `mix_met` matrices), 

- ground_truth.rds

It takes in input source data (script 01.1_src_XXX.Rmd).

# Read source data 

```{r, results="verbatim"}
src = readRDS(paste0("../",source_file))

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met
rm(src)
gc()
```

# Proportion matrix simulation

An example of generative model for 5 cell types, in pancreatic adenocarcinoma.

General distribution :

    normal 15%
    fibro 45%,
    immune 10%,
    cancer 30% (divided between basal and classic)
    
Subdivision of cancer types among cancer cells.

-> in 1/3 samples : 50% type basal, 50% type classic

-> in 1/3 samples : 25% type basal, 75% type classic

-> in 1/3 samples : 10% type basal, 90% type classic



## Function 

```{r, results="verbatim"}
simu_proportion_pancreas = function(n, alpha = c(1.5, 4.5, 1, 3)){
  tmp_mix_Lvar = t(gtools::rdirichlet(n = n, alpha = alpha))
  tumor_rep = cbind(t(gtools::rdirichlet(n = n/3, alpha = c(0.5, 0.5))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.75, 0.25))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.9, 0.1))))
  mix_Lvar = matrix(nrow = 5, ncol = n)
  mix_Lvar [1:3, ] = tmp_mix_Lvar[1:3, ]
  mix_Lvar[4, ] = tumor_rep[1, ]*tmp_mix_Lvar[4, ]
  mix_Lvar[5, ] = tumor_rep[2, ]*tmp_mix_Lvar[4, ]

  return(mix_Lvar)
}
```

## Proportion matrix generation

Below code to generate a list of n number of simulation (length.out)

```{r, results="verbatim", eval=FALSE}

set.seed(10)

ground_truth_list <- lapply(
  X   = seq_len(length.out = 5 ) # length of X = number of simulation instances
  # X   = seq_len(length.out = 1 ) # length of X = number of simulation instances
  , FUN = function( i ) {
    proportion = simu_proportion_pancreas(n = 30) # n -> number of samples
    # rownames(proportion) = c("epi", "fib", "immun", "classic", "basal")
    rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
    return(proportion)
  }
)

saveRDS(ground_truth_list, paste0("../",groundtruth_files))
```

Below code to generate only 1 simulation 

```{r, results="verbatim"}
set.seed(10)

nb_samples = 60

proportion = simu_proportion_pancreas(n = nb_samples) # n -> number of samples
rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
```

## Ground truth - split {-}

```{r}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```

  
# Simulation of DNAm and RNA mixes from convolution Mix_met Mix_RNA

The following function adds noise to the simulation to make it harder to deconvolve.

In order to make the deconvolution more difficult, here the added noise is randomly chosen from distribution with different parameters for different randomly selected subset of genes/met probes. 

Laplace or chi2 law can be used. Issue of chi2 is that it only add positive values. 

```{r, results="verbatim"}

#separated features (gene or methylation probe) in subsets 

subset_features = function(data, #methylation or rna data 
                        n)  # number of subsets
  {
        set.seed(1230) 
        random_order <- sample(nrow(data))
        shuffled_features <- data[random_order, ]
        features = rownames(shuffled_features) 
        
        #groups are separated 
        idx = sample(1:length(features), n-1) # get separation idx 
        idx = append(x = idx,c(1,length(features))) # and Add the first and last idx
        order = order(idx)
        sorted_idx <- idx[order]
        subset_idx_list = list()
        for (i in 1:n) {
          sub_idx = sorted_idx[i]:sorted_idx[i+1]
          subset_idx_list[[i]] = sub_idx
        }
        features_mapped <- lapply(subset_idx_list, function(sublist) features[sublist])
        return(features_mapped) 
  }


add_noise_chi2 = function(features,
                     data,
                     df = 1, # for chi2
                     ncp= 0, # for chi2
                     val_min = 0,
                     val_max
                     )
  { 
  # set.seed(2)

  noise = rchisq(dim(data)[2], df = df, ncp = ncp) - df #generate noise for each cell type = columns, centered around 0 
  noise
  feature_indices <- match(features, rownames(data))
  datam = data # create a modified data to then comprare to original data in case of noise being out of bounds
  datam[feature_indices, ] <- datam[feature_indices, ] + noise
  datam[datam < val_min] = data[datam < val_min] # if the new value with noise is outof bound compared to val_min and val_max it is replaced by the original value
  datam[datam > val_max] = data[datam > val_max]
  return(datam)
}

 
# add_noise_laplace = function(features,
#                      data,
#                      mean = 0, # for laplace
#                      sd = 1, # for laplace
                     # val_min = 0,
                     # val_max
#                      )
#   { 
  set.seed(20)

#   noise = rchisq(dim(data)[2], df = df, ncp = ncp) #generate noise for each cell type = columns
#   feature_indices <- match(features, rownames(data))
#   # feature_indices = features %in% rownames(data)
  # datam = data # create a modified data to then comprare to original data in case of noise being out of bounds
  # datam[feature_indices, ] <- datam[feature_indices, ] + noise
  # datam[datam < val_min] = data[datam < val_min]
  # datam[datam > val_max] = data[datam > val_max]
#   
#   return(datam)
# }


```

## Simulation Mix_RNA

Below code to generate a list of n number of simulation (length.out)

```{r, results="verbatim", eval=FALSE}

#separated genes in subsets to add noise with following different density distribution, subsets are the same between simulations 
n = 4 # number of feature subset 
subset_genes = subset_features(rna_seq_src,n) 

Mix_RNA <- lapply(
  X   = seq_len(length.out = 5 ) # number of simulations
  , FUN = function( i ) {

RNA_brut =  as.matrix(rna_seq_src) %*%  ground_truth_list[[i]] #convolution of source matrix with proportion matrix
# Add noise different noise to each subset of gene 

 #for n = 4
 genes = subset_genes[[1]]
 RNA_brut = add_noise_chi2(genes, RNA_brut, df = 1, val_min = 0, val_max = 10000)
  genes = subset_genes[[2]]
 RNA_brut = add_noise_chi2(genes, RNA_brut, df = 2, val_min = 0, val_max = 10000)
  genes = subset_genes[[3]]
 RNA_brut = add_noise_chi2(genes, RNA_brut, df = 5, val_min = 0, val_max = 10000)
  genes = subset_genes[[4]]
 RNA_brut = add_noise_chi2(genes, RNA_brut, df = 10, val_min = 0, val_max = 10000)

  }
)


head(Mix_RNA[[1]])
```

Below code to generate only 1 simulation 

```{r, results="verbatim"}
Mix_RNA =  as.matrix(rna_seq_src) %*% proportion #convolution of source matrix with proportion matrix

#separated genes in subsets to add noise with following different density distribution

n = 4 # number of feature subset 
subset_genes = subset_features(Mix_RNA,n)

# Add noise different noise to each subset of gene 
 #for n = 4
 genes = subset_genes[[1]]
 Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 1, val_min = 0, val_max = 10000)
  genes = subset_genes[[2]]
 Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 2, val_min = 0, val_max = 10000)
  genes = subset_genes[[3]]
 Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 5, val_min = 0, val_max = 10000)
  genes = subset_genes[[4]]
 Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 10, val_min = 0, val_max = 10000)
```

## Simulation Mix_met

Below code to generate a list of n number of simulation (length.out)

```{r, results="verbatim", eval=FALSE}
#separated probes in subsets to add noise with following different density distribution
set.seed(20)

n = 4 # number of feature subset 
subset_probes = subset_features(dna_met_src,n)

Mix_met <- lapply(
  X   = seq_len(length.out = 5 )
  , FUN = function( i ) {
     met_brut =  as.matrix(dna_met_src) %*% ground_truth_list[[i]] #convolution of source matrix with proportion matrix
    # Add noise different noise to each subset of gene 
    
     #for n = 4
     probes = subset_probes[[1]]
     met_brut = add_noise_chi2(probes, met_brut, df = 0.1, val_min = 0, val_max = 1)
      probes = subset_probes[[2]]
     met_brut = add_noise_chi2(probes, met_brut, df = 0.05, val_min = 0, val_max = 1)
      probes = subset_probes[[3]]
     met_brut = add_noise_chi2(probes, met_brut, df = 0.02, val_min = 0, val_max = 1)
      probes = subset_probes[[4]]
     met_brut = add_noise_chi2(probes, met_brut, df = 0.06, val_min = 0, val_max = 1)
     
 }
)

head(Mix_met[[1]])
```

Below code to generate only 1 simulation 

```{r, results="verbatim"}
set.seed(20)

Mix_met =  as.matrix(dna_met_src) %*% proportion #convolution of source matrix with proportion matrix

#separated probes in subsets to add noise with following different density distribution

n = 4 # number of feature subset 
subset_probes = subset_features(Mix_met,n)

# Add noise different noise to each subset of gene 

 #for n = 4
 probes = subset_probes[[1]]
 Mix_met = add_noise_chi2(probes, Mix_met, df = 0.1, val_min = 0, val_max = 1)
  probes = subset_probes[[2]]
 Mix_met = add_noise_chi2(probes, Mix_met, df = 0.05, val_min = 0, val_max = 1)
  probes = subset_probes[[3]]
 Mix_met = add_noise_chi2(probes, Mix_met, df = 0.02, val_min = 0, val_max = 1)
  probes = subset_probes[[4]]
 Mix_met = add_noise_chi2(probes, Mix_met, df = 0.06, val_min = 0, val_max = 1)

head(Mix_met)
```

## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=Mix_RNA[,idx_phaseUsers],
                              mix_met=Mix_met[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=Mix_RNA[,-idx_phaseUsers],
                             mix_met=Mix_met[,-idx_phaseUsers])
```

# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```

