  ---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---


```{r, include=(FALSE)}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=TRUE, results="verbatim")
options(knitr.duplicate.label = "allow")
```

# Parameters

Here we set up input and outputs files for the vignette.

```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
```

# Libraries 

```{r library}
library(Seurat)
```

# Objectives

This script generates:

- references_data.rds (a list with `ref_bulkRNA`, `ref_scRNA`, `ref_met`)

It takes as input raw data.


# Reference Data 

COMT40TFZN_D1	Pure Neutrophils (DNA part of RNA/DNA double extraction) \
COMT41TFZN_D1	Pure Endothelial cells (DNA part of RNA/DNA double extraction) 
COMT42TFZN_D1	Pure Macrophages (DNA part of RNA/DNA double extraction) \
COMT43TFZC_D1	Pure Cancer Classical (DNA part of RNA/DNA double extraction) \
COMT44TFZN_D1	Pure B cells (DNA part of RNA/DNA double extraction) \
COMT45TFZN_D1	Pure CD8 T Cells (DNA part of RNA/DNA double extraction) \
COMT46TFZB_D1	Pure Cancer Basal (DNA part of RNA/DNA double extraction) \
COMT47TFZN_D1	Pure CD4 T cells (DNA part of RNA/DNA double extraction) \
COMT48TFZT_D1	Pure Fibroblast (DNA part of RNA/DNA double extraction) \

## Pure cell lines bulk RNA seq and DNAmet from cometh lot 1 

```{r}
#Data cometh lot 1 : 30 in vitro mixes with DNAm and RNA-seq + 9 pure cell lines
#rna = read.table("~/projects/datashare/cometh_lot1/transcriptome/gene_counts_filtered.txt") not normalized
rna = read.table("~/projects/datashare/cometh_lot1/transcriptome/gene_counts_normalised.txt")

# colnames(rna[[1]])
met = readRDS("~/projects/datashare/cometh_lot1/methylation/test_data_met_new.rds")

# colnames(met[[1]])
# the 9 last samples are pure samples. 
pure_names_dict = c(COMT40TFZN="neutrophils"      , 
                    COMT41TFZN="endothelial_cells",
                    COMT42TFZN="macrophages"      ,
                    COMT43TFZC="tum_classical"    ,
                    COMT44TFZN="b_cells"          ,
                    COMT45TFZN="cd8_t_cells"      ,
                    COMT46TFZB="tum_basal"        ,
                    COMT47TFZN="cd4_t_cells"      ,
                    COMT48TFZT="fibroblasts"      )


pure_rna_seq = data.frame(rna[, names(pure_names_dict)]) #get pure cell types
pure_dna_met = data.frame(met[, names(pure_names_dict)]) #get pure cell types
# all(colnames(pure_rna_seq) == colnames(pure_dna_met)) #check IDs 


colnames(pure_rna_seq) = pure_names_dict[colnames(pure_rna_seq)]
colnames(pure_dna_met) = pure_names_dict[colnames(pure_dna_met)]

#  change names of cell type to accommodate ground_truth_list

#remove NA values in rna dataset
# Not like that:
# idx_na = which(is.na(rowMeans(pure_rna_seq))) 
# pure_rna_seq = pure_rna_seq[-idx_na,] # remove NA 
# Better like that: 
pure_rna_seq = pure_rna_seq[apply(!is.na(pure_rna_seq), 1, all),] # remove NA 
dim(pure_rna_seq)
dim(pure_dna_met)

current_rna = cbind(endo = pure_rna_seq$endothelial_cells, #change later for epithelial cells 
                        fibro = pure_rna_seq$fibroblasts,
                        immune = rowMeans(pure_rna_seq[, c("cd4_t_cells", "b_cells", "macrophages", "cd8_t_cells", "neutrophils")]), 
                        classic = pure_rna_seq$tum_classical,
                        basal = pure_rna_seq$tum_basal
                        )
#get back gene names 
rownames(current_rna) = rownames(pure_rna_seq)

current_met = cbind(endo = pure_dna_met$endothelial_cells, #change later for epithelial cells 
                        fibro = pure_dna_met$fibroblasts,
                        immune = rowMeans(pure_dna_met[, c("cd4_t_cells", "b_cells", "macrophages", "cd8_t_cells", "neutrophils")]), 
                        classic = pure_dna_met$tum_classical,
                        basal = pure_dna_met$tum_basal
                        )
rownames(current_met) = rownames(pure_dna_met)

head(current_rna)

head(current_met)
```

## Single cell RNAseq

### All cells

```{r}
# Labelled single cells from previous deconv_pdac analysis (from Peng dataset)
# mkdir -p ~/projects/datashare/pancreas_deconv_data/results/20210624/02.1_output/
# rsync -auvP dahu.ciment:/bettik/richamag/projects/pancreas_deconv_data/results/20210624/02.1_output/peng_pool_labelled.rds ~/projects/datashare/pancreas_deconv_
pool_pen = readRDS("~/projects/pancreas_deconv_data/results/20210624/02.1_output/peng_pool_labelled.rds")
table(pool_pen@meta.data$label_full)

# get all indices 
idx_endo <- which(pool_pen@meta.data$label_full == "Endothelial cells")
idx_immune <- which(pool_pen@meta.data$label_full %in% c("B cells","Mono.Macro","T cells","Dendritic cells"))
idx_fibro <- which(pool_pen@meta.data$label_full == "Fibro/Stellate")
idx_classic <- which(pool_pen@meta.data$label_full == "Classic")
idx_basal <- which(pool_pen@meta.data$label_full == "Basal")

#create a new metadata column with hadaca3_labels to accomodate ground_truth list

pool_pen@meta.data[idx_endo,"hadaca3"] <- "endo" # change later for epi 
pool_pen@meta.data[idx_immune,"hadaca3"] <- "immune"
pool_pen@meta.data[idx_fibro,"hadaca3"] <- "fibro"
pool_pen@meta.data[idx_classic,"hadaca3"] <- "classic"
pool_pen@meta.data[idx_basal,"hadaca3"] <- "basal"

# create reference data (T matrix) from peng data 

cell_types <- c("immune","endo","fibro","basal","classic")

subset_hadaca <- subset(pool_pen, subset = hadaca3 %in% cell_types) #remove all cells without cell type assignment in hadaca3 metadata field

counts <- subset_hadaca@assays$RNA@counts #type dgCMatrix (sparse matrix)
# counts_matrice <- as.matrix(counts) # to converte sparse matrix into regular matrix but would be 3.5 Gb
metadata <- as.matrix(subset_hadaca@meta.data$hadaca3)

#rename col and row of metadata matrix
rownames(metadata) <- rownames(subset_hadaca@meta.data)
colnames(metadata) <- "cell_types"

sc_rna_ref <- list(counts=counts, metadata=metadata)
```

### 1000 cells sample
This next chunk is set with "eval=FALSE", it will not be compiled. To run it, remove "eval=FALSE" and add it to the previous chunk

```{r, results="verbatim", eval=FALSE}
# Labelled single cells from previous deconv_pdac analysis (from Peng dataset)
pool_pen = readRDS("~/projects/pancreas_deconv_data/results/20210624/02.1_output/peng_pool_labelled.rds")
table(pool_pen@meta.data$label_full)

  
# get all indices 
idx_endo <- which(pool_pen@meta.data$label_full == "Endothelial cells")
idx_immune <- which(pool_pen@meta.data$label_full %in% c("B cells","Mono.Macro","T cells","Dendritic cells"))
idx_fibro <- which(pool_pen@meta.data$label_full == "Fibro/Stellate")
idx_classic <- which(pool_pen@meta.data$label_full == "Classic")
idx_basal <- which(pool_pen@meta.data$label_full == "Basal")

#create a new metadata column with hadaca3 labels to accommodate ground_truth list

pool_pen@meta.data[idx_endo,"hadaca3"] <- "endo" # change later for epi 
pool_pen@meta.data[idx_immune,"hadaca3"] <- "immune"
pool_pen@meta.data[idx_fibro,"hadaca3"] <- "fibro"
pool_pen@meta.data[idx_classic,"hadaca3"] <- "classic"
pool_pen@meta.data[idx_basal,"hadaca3"] <- "basal"

#compute cell type proportion in peng to keep it after sampling
tot_cell_hadaca3 <- length(c(idx_endo,idx_immune,idx_fibro,idx_classic,idx_basal))
prop_endo <- length(idx_endo)/tot_cell_hadaca3
prop_immune <- length(idx_immune)/tot_cell_hadaca3
prop_fibro <- length(idx_fibro)/tot_cell_hadaca3
prop_classic <- length(idx_classic)/tot_cell_hadaca3
prop_basal <- length(idx_basal)/tot_cell_hadaca3

#sample each cell type in proportion for a 1000 cells
sample_endo <- sample(idx_endo, prop_endo*1000)
sample_immune <- sample(idx_immune, prop_immune*1000)
sample_fibro <- sample(idx_fibro, prop_fibro*1000)
sample_classic <- sample(idx_classic, prop_classic*1000)
sample_basal <- sample(idx_basal, prop_basal*1000)

# create a subset with the sample index
subset_hadaca_1000 <- subset(pool_pen, cells = c(sample_basal,sample_classic,sample_endo, sample_fibro, sample_immune))

metadata <- as.matrix(subset_hadaca@meta.data$hadaca3)

#rename col and row of metadata matrix
rownames(metadata) <- rownames(subset_hadaca_1000@meta.data)
cell_types <- c("immune","endo","fibro","basal","classic")
colnames(metadata) <- "cell_types"

sc_rna_ref <- list(counts=counts,metadata=metadata)
```

### Save into list

```{r, results="verbatim"}
#T matrices
source_data <- list(src_bulkRNA=current_rna, src_scRNA=sc_rna_ref, src_met=current_met)

saveRDS(source_data, paste0(source_file))
```

# Session Information

```{r, results="verbatim"}
sessionInfo()
```
