---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)
library(Seurat)
```


```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes1_insilicodirichletNoDep_pdac.rds","mixes2_insilicodirichletNoDep_pdac.rds")}
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth1_insilicodirichletNoDep_pdac.rds","groundtruth2_insilicodirichletNoDep_pdac.rds")}
```

# Objectives

This script generates :

- mixes_data.rds (a list with `mix_rna` and `mix_met` matrices), 

- ground_truth.rds

It takes as input source data (script 01.1_src_XXX.Rmd).


# Read source data 

```{r, results="verbatim"}
src = readRDS(paste0("../",source_file))

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met

rm(src)
gc()
```

# Proportion matrix simulation

An example of generative model for 5 cell types, in pancreatic adenocarcinoma.

General distribution :

    normal 15%
    fibro 45%,
    immune 10%,
    cancer 30% (divided between basal and classical subtypes)
    
Subdivision of cancer types among cancer cells.

-> in 1/3 of the samples : 50% type basal, 50% type classical

-> in 1/3 of the samples : 25% type basal, 75% type classical

-> in 1/3 of the samples : 10% type basal, 90% type classical


## Proportion matrix generation


Bellow code to generate only 1 simulation 

```{r prop_mat_generation, results="verbatim"}
# set.seed(1)

nb_samples = 60



# alpha_prop1 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45, 0.1, 0.5*0.3, 0.5*0.3), # 0.15, 0.15
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop1 = c(5.162500, 15.487500, 3.441667, 5.162500, 5.162500)

# alpha_prop2 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.75*0.3, 0.25*0.3), # 0.225, 0.075
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop2 = c(5.162500, 15.487500, 3.441667, 7.743750, 2.581250)

# alpha_prop3 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.9*0.3, 0.1*0.3), # 0.27, 0.03
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop3 = c(5.162500, 15.487500, 3.441667, 9.292500, 1.032500)


n = floor(nb_samples/3)
proportion = t(rbind(
  gtools::rdirichlet(n = n + (nb_samples - 3*n), alpha = alpha_prop1),
  gtools::rdirichlet(n = n, alpha = alpha_prop2),
  gtools::rdirichlet(n = n, alpha = alpha_prop3)
))

rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal")

rm(alpha_prop1, alpha_prop2, alpha_prop3, n)
```


## Ground truth - split {-}

```{r}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```


# Simulation

Here we do a convolution of source data using the proportion matrix to simulate data. In order to make the deconvolution more difficult, the added noise is randomly chosen from distribution with different parameters for different randomly selected subset of genes/met probes. 



The following function adds noise to the simulation to make the deconvolution harder and the simulation more realistic.

```{r, results="verbatim"}

#separated features (gene or methylation probe) in subsets 
subset_features = function(data, #methylation or rna data 
                           n)  # number of subsets
  {
    # set.seed(123) 
    random_order <- sample(nrow(data))
    shuffled_features <- data[random_order, ]
    features = rownames(shuffled_features) 
    
    #groups are separated 
    idx = sample(1:length(features), n-1) # get separation idx 
    idx = append(x = idx,c(1,length(features))) # and Add the first and last idx
    order = order(idx)
    sorted_idx <- idx[order]
    subset_idx_list = list()
    for (i in 1:n) {
      sub_idx = sorted_idx[i]:sorted_idx[i+1]
      subset_idx_list[[i]] = sub_idx
    }
    features_mapped <- lapply(subset_idx_list, function(sublist) features[sublist])
    return(features_mapped) 
  }


# function to add noise on D_rna matrix
add_noise_chi2 = function(features,
                          data,
                          df = 1, # for chi2
                          ncp= 0, # for chi2
                          val_min = 0,
                          val_max) { 
  
  feature_indices <- match(features, rownames(data))
  noise = t(replicate(n = length(feature_indices), expr = (rchisq(dim(data)[2], df = df, ncp = ncp) - df), simplify = "array"))
  
  datam = data # create a modified data to then comprare to original data in case of noise being out of bounds
  datam[feature_indices,] <- datam[feature_indices,] + noise
  # if the new value with noise is outof bound compared to val_min and val_max it is replaced by the original value + a minor variation
  datam[datam < val_min] = jitter(data[datam < val_min], amount = 1e-6)
  datam[datam > val_max] = jitter(data[datam > val_max], amount = 1e-6)
  # double check for the minor variation
  datam[datam < val_min] = data[datam < val_min]
  datam[datam > val_max] = data[datam > val_max]
  return(datam)
}


# function to add noise on D_met matrix
add_noise_norm = function(features,
                          data, 
                          mean = 0, 
                          sd = 0.05, 
                          val_min = 0, 
                          val_max = 1) {  # val min et val max make sure the value stays biologicaly coherent ([0:1] for met and [0:...] for rna)
  
  feature_indices <- match(features, rownames(data))
  noise = t(replicate(n = length(feature_indices), expr = rnorm(dim(data)[2], mean = mean, sd = sd), simplify = "array")) # need to transpose the replicate 
  
  datam = data # create a modified data to then comprare to original data in case of noise being out of bounds
  datam[feature_indices,] = datam[feature_indices,] + noise
  # if the new value with noise is outof bound compared to val_min and val_max it is replaced by the original value + a minor variation
  datam[datam < val_min] = jitter(data[datam < val_min], amount = 1e-6)
  datam[datam > val_max] = jitter(data[datam > val_max], amount = 1e-6)
  # double check for the minor variation
  datam[datam < val_min] = data[datam < val_min]
  datam[datam > val_max] = data[datam > val_max]
  return(datam)
}
```

### Simulation Mix_RNA


Below is the code to generate only 1 simulation 

```{r, results="verbatim"}
# set.seed(2)
Mix_RNA =  as.matrix(rna_seq_src) %*% proportion #convolution of source matrix with proportion matrix

#separated genes in subsets to add noise with following different density distribution

n = 4 # number of feature subset 
subset_genes = subset_features(Mix_RNA,n)

# Add noise different noise to each subset of gene 
#for n = 4
genes = subset_genes[[1]]
Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 1, val_min = 0, val_max = 10000)
genes = subset_genes[[2]]
Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 2, val_min = 0, val_max = 10000)
genes = subset_genes[[3]]
Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 5, val_min = 0, val_max = 10000)
genes = subset_genes[[4]]
Mix_RNA = add_noise_chi2(genes, Mix_RNA, df = 10, val_min = 0, val_max = 10000)

# tmptmp = cor(t(Mix_RNA - as.matrix(rna_seq_src) %*% proportion))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of true error simulated by Chi2\nwithout explicit dependence (rna)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))
```

### Simulation Mix_met

Below is the code to generate only 1 simulation 

```{r, results="verbatim"}
# set.seed(2)
Mix_met =  as.matrix(dna_met_src) %*% proportion #convolution of source matrix with proportion matrix

n = 4 # number of feature subset 
subset_probes = subset_features(Mix_met,n)

# Add noise different noise to each subset of gene 
#for n = 4
probes = subset_probes[[1]]
Mix_met = add_noise_norm(probes, Mix_met, mean = 0, sd = 0.01, val_min = 0, val_max = 1)
probes = subset_probes[[2]]
Mix_met = add_noise_norm(probes, Mix_met, mean = 0, sd = 0.05, val_min = 0, val_max = 1)
probes = subset_probes[[3]]
Mix_met = add_noise_norm(probes, Mix_met, mean = 0, sd = 0.1, val_min = 0, val_max = 1)
probes = subset_probes[[4]]
Mix_met = add_noise_norm(probes, Mix_met, mean = 0, sd = 0.2, val_min = 0, val_max = 1)


# tmp_sample = sample(1:nrow(Mix_met), 20000, replace = FALSE)
# tmptmp = cor(t(Mix_met[tmp_sample,] - as.matrix(dna_met_src)[tmp_sample,] %*% proportion))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of true error simulated by norm\nwithout explicit  dependence (met)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))
```


## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=Mix_RNA[,idx_phaseUsers],
                              mix_met=Mix_met[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=Mix_RNA[,-idx_phaseUsers],
                             mix_met=Mix_met[,-idx_phaseUsers])
```

# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```
