---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe, Hugo Barbot and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r pckg, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)

if ( !( "copula" %in% installed.packages() ) ) {
  install.packages("copula", repos = "http://cran.us.r-project.org")
}
library(copula)
```


```{r params}
if (!exists("source_file")) {source_file = "01_references/source_pdac.rds"} 
if (!exists("mixes_files")) {mixes_files = c("01_mixes/mixes_insilicodirichletCopule1_pdac.rds","01_mixes/mixes_insilicodirichletCopule2_pdac.rds")}
if (!exists("groundtruth_files")) {groundtruth_files = c("01_groundtruth/groundtruth_insilicodirichletCopule1_pdac.rds","01_groundtruth/groundtruth_insilicodirichletCopule2_pdac.rds")} 
```

# Objectives

This script generates :

- mixes_data.rds (a list with `mix_rna` and `mix_met` matrices), 

- ground_truth.rds

It takes in input source data (script 01.1_src_XXX.Rmd).

# Read source data 

```{r read_source_file, results="verbatim"}
src = readRDS(paste0("../",source_file))

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met
rm(src)
gc()
```

# Proportion matrix simulation

An example of generative model for 5 cell types, in pancreatic adenocarinoma.

General distribution :

    normal 15%
    fibro 45%,
    immune 10%,
    cancer 30% (divided between basal and classic)
    
Subdivision of cancer types among cancer cells.

-> in 1/3 samples : 50% type basal, 50% type classic

-> in 1/3 samples : 25% type basal, 75% type classic

-> in 1/3samples : 10% type basal, 90% type classic


## Proportion matrix generation

Below code to generate only 1 simulation 

```{r prop_mat_generation, results="verbatim"}
# set.seed(1)
nb_samples = 60

# alpha_prop1 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45, 0.1, 0.5*0.3, 0.5*0.3), # 0.15, 0.15
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop1 = c(5.162500, 15.487500, 3.441667, 5.162500, 5.162500)

# alpha_prop2 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.75*0.3, 0.25*0.3), # 0.225, 0.075
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop2 = c(5.162500, 15.487500, 3.441667, 7.743750, 2.581250)

# alpha_prop3 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.9*0.3, 0.1*0.3), # 0.27, 0.03
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop3 = c(5.162500, 15.487500, 3.441667, 9.292500, 1.032500)


n = floor(nb_samples/3)
proportion = t(rbind(
  gtools::rdirichlet(n = n + (nb_samples - 3*n), alpha = alpha_prop1),
  gtools::rdirichlet(n = n, alpha = alpha_prop2),
  gtools::rdirichlet(n = n, alpha = alpha_prop3)
))

rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal")

rm(alpha_prop1, alpha_prop2, alpha_prop3, n)
```

```{r, results="verbatim", eval=FALSE}
set.seed(10)

### Deprecated and can have only a number of samples divisible by 3

simu_proportion_pancreas = function(n, alpha = c(1.5, 4.5, 1, 3)){
  tmp_mix_Lvar = t(gtools::rdirichlet(n = n, alpha = alpha))
  tumor_rep = cbind(t(gtools::rdirichlet(n = n/3, alpha = c(0.5, 0.5))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.75, 0.25))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.9, 0.1))))
  mix_Lvar = matrix(nrow = 5, ncol = n)
  mix_Lvar [1:3, ] = tmp_mix_Lvar[1:3, ]
  mix_Lvar[4, ] = tumor_rep[1, ]*tmp_mix_Lvar[4, ]
  mix_Lvar[5, ] = tumor_rep[2, ]*tmp_mix_Lvar[4, ]

  return(mix_Lvar)
}

nb_samples = 60

proportion = simu_proportion_pancreas(n = nb_samples) # n -> number of samples
# rownames(proportion) = c("epi", "fib", "immun", "classic", "basal")
rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
```

## Ground truth - split {-}

```{r prop_mat_saving, results="verbatim"}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1])) 
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```


# Need of true epsilon error

We based this simulation on the empiric copula of true epsilon error $\varepsilon_{true} = Y_{bulk} - X_{src} \times \beta_{true}$. So we copy script from `01.2_dataset_invitro`.


```{r}
ground_truth = readRDS("~/projects/datashare/cometh_lot1/test_solution.rds")[[1]][,1:30]

colnames(rna_seq_src)
rownames(ground_truth)
mask_ref = c("immune","basal","classic","immune","immune","endo","fibro","immune","immune")

ground_truth_with_mask = t(sapply(colnames(rna_seq_src), function(celltype)
  if (sum(mask_ref==celltype)==1) {
    ground_truth[mask_ref==celltype,]
  } else {colSums(ground_truth[mask_ref==celltype,])}
))
```

```{r}
# Loading the DESeq normalised RNA matrix
rna_measure = read.table("~/projects/datashare/cometh_lot1/transcriptome/gene_counts_normalised.txt")
rna_measure = as.matrix(rna_measure)
# Loading the filtered-normalised DNAm matrix
met_measure = readRDS("~/projects/datashare/cometh_lot1/methylation/test_data_met_new.rds")

rna_measure = rna_measure[,colnames(ground_truth_with_mask)]
met_measure = met_measure[,colnames(ground_truth_with_mask)]
```

# Simulation of DNAm and RNA mixes from convolution Mix_met Mix_RNA

The following function adds noise to the simulation to make is harder to deconvolve, more life like.

In order to harder the deconvolution, here the added noise is randomly generated with a empirical copula from the noise of in vitro dataset and chosen from distribution with different parameters for each genes/met probes. 



## Simulation Mix_RNA


```{r Estimate_RNA_ground_truth_epsilon_and_empirical_copula}
ground_truth = ground_truth_with_mask
expr = rna_measure
N = ncol(ground_truth)

epsilon_in_vitro_rna = expr - rna_seq_src %*% ground_truth

mean_epsilon_in_vitro_rna = rowMeans(epsilon_in_vitro_rna)
c_epsilon_in_vitro_rna = epsilon_in_vitro_rna - tcrossprod(mean_epsilon_in_vitro_rna, rep(1,N))  # center
sd_epsilon_in_vitro_rna = sqrt(rowMeans(c_epsilon_in_vitro_rna^2))
sc_epsilon_in_vitro_rna = c_epsilon_in_vitro_rna/tcrossprod(sd_epsilon_in_vitro_rna,rep(1,N))    # scale

t_eps_expr = t(sc_epsilon_in_vitro_rna)
t_eps_expr_pseudo_obs = pobs(t_eps_expr)
rownames(t_eps_expr_pseudo_obs) = NULL

fitted_emp_copula_expr = empCopula(t_eps_expr_pseudo_obs)

rm(expr, mean_epsilon_in_vitro_rna, c_epsilon_in_vitro_rna, sd_epsilon_in_vitro_rna, sc_epsilon_in_vitro_rna, t_eps_expr, t_eps_expr_pseudo_obs) # ground_truth, N
```


```{r RNA_copula_noise_with_marginals, results="verbatim"}
noise_rna_aggr_without_marginals = rCopula(nb_samples, fitted_emp_copula_expr)
noise_rna_aggr_without_marginals = t(noise_rna_aggr_without_marginals)

# tmptmp = cor(t(noise_rna_aggr_without_marginals))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of quantile simulated by empiraical copula (rna)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

stopifnot(identical(rownames(noise_rna_aggr_without_marginals), rownames(epsilon_in_vitro_rna)))
noise_rna_aggr = sapply(1:nrow(noise_rna_aggr_without_marginals), 
  function(ligne_i){
    
    if (min(epsilon_in_vitro_rna[ligne_i,])<0) {
      tmp_epsilon_in_vitro_rna = epsilon_in_vitro_rna[ligne_i,] - min(epsilon_in_vitro_rna[ligne_i,])
    } else {
      tmp_epsilon_in_vitro_rna = epsilon_in_vitro_rna[ligne_i,]
    }
    
    n_data = length(tmp_epsilon_in_vitro_rna)
    
    mean_data = mean(tmp_epsilon_in_vitro_rna)
    var_data = (n_data - 1)/n_data * var(tmp_epsilon_in_vitro_rna) # debiased variance
    size_estim = mean_data^2 / (var_data - mean_data)
    
    if (size_estim <= 0) { # always with very low data
      size_estim = 0.5
    }
    
    # hist(tmp_epsilon_in_vitro_rna, freq = FALSE, nclass = 100)
    # lines(seq(0,max(tmp_epsilon_in_vitro_rna),1), stats::dnbinom(seq(0,max(tmp_epsilon_in_vitro_rna),1), size = size_estim, mu = mean_data))

    
    eps_gene_i = stats::qnbinom(noise_rna_aggr_without_marginals[ligne_i,], size = size_estim, mu = mean_data)
    
    if (min(epsilon_in_vitro_rna[ligne_i,])<0) {
      eps_gene_i = eps_gene_i - mean(eps_gene_i) + mean(epsilon_in_vitro_rna[ligne_i,]) # pour remettre la bonne moyenne
    }
    
    return(eps_gene_i)
  }
)
noise_rna_aggr = t(noise_rna_aggr)
rownames(noise_rna_aggr) = rownames(noise_rna_aggr_without_marginals)

# i = 45 ou 115 ou "MIR6723"
# i = sample(1:nrow(noise_rna_aggr_without_marginals), 1)
# 
# par(mfrow = c(2,1))
# hist(epsilon_in_vitro_rna[i,])
# hist(noise_rna_aggr[i,])
# 
# 
# j = sample(1:nrow(noise_rna_aggr_without_marginals), 1)
# 
# par(mfrow = c(1,2))
# plot(epsilon_in_vitro_rna[j,], epsilon_in_vitro_rna[i,], pch = 20)
# plot(noise_rna_aggr[j,], noise_rna_aggr[i,], pch = 20)
# par(mfrow = c(1,1))

rm(epsilon_in_vitro_rna, noise_rna_aggr_without_marginals, fitted_emp_copula_expr)
```


```{r adding_Mix_RNA_and_noise, results="verbatim"}
Mix_RNA = as.matrix(rna_seq_src) %*% proportion #convolution of source 

Mix_RNA = Mix_RNA + noise_rna_aggr

# sum(apply(Mix_RNA, 1, \(gene_i){any(is.na(gene_i))}))
which_gene_neg = apply(Mix_RNA, 1, \(gene_i){any(gene_i<0)})
which_gene_neg = names(which_gene_neg[which(which_gene_neg)])

for (gene_i in which_gene_neg) {
  # ddl_gene_i = sd_epsilon_in_vitro_rna[gene_i]
  Mix_RNA[gene_i,] = Mix_RNA[gene_i,] - min(Mix_RNA[gene_i,]) + runif(nb_samples, min = 0, max = 1e-3)
}

# tmp = Mix_RNA - as.matrix(rna_seq_src) %*% proportion # noise_rna_aggr
# tmptmp = cor(t(tmp))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of residual correlations for simulation by emfa (rna)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

rm(noise_rna_aggr, which_gene_neg)
gc()
```


## Simulation Mix_met


```{r Estimate_met_ground_truth_epsilon_and_empirical_copula}
# ground_truth = cbind(readRDS("groundtruth1_invitro_pdac.rds"), readRDS("groundtruth2_invitro_pdac.rds"))
meth = met_measure
# N = ncol(ground_truth)

epsilon_in_vitro_meth = meth - dna_met_src %*% ground_truth

mean_epsilon_in_vitro_meth = rowMeans(epsilon_in_vitro_meth)
c_epsilon_in_vitro_meth = epsilon_in_vitro_meth - tcrossprod(mean_epsilon_in_vitro_meth, rep(1,N))  # center
sd_epsilon_in_vitro_meth = sqrt(rowMeans(c_epsilon_in_vitro_meth^2))
sc_epsilon_in_vitro_meth = c_epsilon_in_vitro_meth/tcrossprod(sd_epsilon_in_vitro_meth,rep(1,N))    # scale

t_eps_meth = t(sc_epsilon_in_vitro_meth)
t_eps_meth_pseudo_obs = pobs(t_eps_meth)
rownames(t_eps_meth_pseudo_obs) = NULL

fitted_emp_copula_meth = empCopula(t_eps_meth_pseudo_obs)

rm(ground_truth, meth, N, mean_epsilon_in_vitro_meth, c_epsilon_in_vitro_meth, sd_epsilon_in_vitro_meth, sc_epsilon_in_vitro_meth, t_eps_meth, t_eps_meth_pseudo_obs)
```


```{r met_copula_noise_with_marginals, results="verbatim"}
noise_met_aggr_without_marginals = rCopula(nb_samples, fitted_emp_copula_meth)
noise_met_aggr_without_marginals = t(noise_met_aggr_without_marginals)

# tmptmp = cor(t(noise_met_aggr_without_marginals))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of quantile simulated by empiraical copula (met)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

stopifnot(identical(rownames(noise_met_aggr_without_marginals), rownames(epsilon_in_vitro_meth)))
noise_met_aggr = sapply(1:nrow(noise_met_aggr_without_marginals), 
  function(ligne_i){
    
    if (min(epsilon_in_vitro_meth[ligne_i,])<0) {
      tmp_epsilon_in_vitro_met = epsilon_in_vitro_meth[ligne_i,] - min(epsilon_in_vitro_meth[ligne_i,])
    } else {
      tmp_epsilon_in_vitro_met = epsilon_in_vitro_meth[ligne_i,]
    }

    param_tmp = mean(tmp_epsilon_in_vitro_met) * (1 - mean(tmp_epsilon_in_vitro_met)) / var(tmp_epsilon_in_vitro_met)
    a = mean(tmp_epsilon_in_vitro_met) * (param_tmp - 1)              # = shape1
    b = (1 - mean(tmp_epsilon_in_vitro_met)) * (param_tmp - 1)        # = shape2
    
    eps_probe_i = qbeta(noise_met_aggr_without_marginals[ligne_i,], shape1 = a, shape2 = b)
    
    if (min(epsilon_in_vitro_meth[ligne_i,])<0) {
      eps_probe_i = eps_probe_i - mean(eps_probe_i) + mean(epsilon_in_vitro_meth[ligne_i,]) # pour remettre la bonne moyenne
    }
    
    return(eps_probe_i)
  }
)
noise_met_aggr = t(noise_met_aggr)
rownames(noise_met_aggr) = rownames(noise_met_aggr_without_marginals)

# i = 45 ou 115
# i = sample(1:nrow(noise_met_aggr_without_marginals), 1)
# 
# par(mfrow = c(2,1))
# plot(epsilon_in_vitro_meth[i,], pch = 20)
# plot(noise_met_aggr[i,], pch = 20)
# 
# 
# j = sample(1:nrow(noise_met_aggr_without_marginals), 1)
# 
# par(mfrow = c(1,2))
# plot(epsilon_in_vitro_meth[j,], epsilon_in_vitro_meth[i,], pch = 20)
# plot(noise_met_aggr[j,], noise_met_aggr[i,], pch = 20)
# par(mfrow = c(1,1))

rm(epsilon_in_vitro_meth, noise_met_aggr_without_marginals, fitted_emp_copula_meth)
```


```{r adding_Mix_met_and_noise_and_removing_neg_value, results="verbatim"}
Mix_met = as.matrix(dna_met_src) %*% proportion #convolution of reference matrix with proportion matrix

Mix_met = Mix_met + noise_met_aggr
Mix_met[Mix_met<0] = runif(sum(Mix_met<0), min = 0, max = 1e-03)
Mix_met[Mix_met>1] = runif(sum(Mix_met>1), min = 1-1e-03, max = 1)

# tmp_sample = sample(1:nrow(Mix_met), 20000, replace = FALSE)
# tmp = Mix_met[tmp_sample,] - as.matrix(dna_met_src[tmp_sample,]) %*% proportion
# tmptmp = cor(t(tmp))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# 
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of residual correlations for simulation by empirical copula (met)",
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

rm(noise_met_aggr)
gc()
```

## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=Mix_RNA[,idx_phaseUsers],
                              mix_met=Mix_met[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=Mix_RNA[,-idx_phaseUsers],
                             mix_met=Mix_met[,-idx_phaseUsers])
```

# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```

