---
title: "HADACA3 - Generate data for data challenge hadaca"
author: "Lucie Lamothe, Hugo Barbot and Magali Richard"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---
# Libraries 

```{r, include=(FALSE)}
knitr::opts_chunk$set(echo = TRUE)
# library(Seurat)
```


```{r params}
if (!exists("source_file")) {source_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes_insilicodirichletFixedfactor1_pdac.rds","mixes_insilicodirichletFixedfactor2_pdac.rds")} 
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth_insilicodirichletFixedfactor1_pdac.rds","groundtruth_insilicodirichletFixedfactor2_pdac.rds")} 
```

# Objectives

This script generates :

- mixes_data.rds (a list with `mix_rna` and `mix_met` matrices), 

- ground_truth.rds

It takes in input source data (script 01.1_src_XXX.Rmd).

# Read source data 

```{r, results="verbatim"}

src = readRDS(paste0("../",source_file))

rna_seq_src = src$src_bulkRNA
dna_met_src = src$src_met
rm(src)
gc()

```

# Proportion matrix simulation

An example of generative model for 5 cell types, in pancreatic adenocarinoma.

General distribution :

    normal 15%
    fibro 45%,
    immune 10%,
    cancer 30% (divided between basal and classic)
    
Subdivision of cancer types among cancer cells.

-> in 1/3 samples : 50% type basal, 50% type classic

-> in 1/3 samples : 25% type basal, 75% type classic

-> in 1/3 samples : 10% type basal, 90% type classic




## Proportion matrix generation

Below code to generate only 1 simulation 

```{r prop_mat_generation, results="verbatim"}
# set.seed(1)

nb_samples = 60



# alpha_prop1 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45, 0.1, 0.5*0.3, 0.5*0.3), # 0.15, 0.15
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop1 = c(5.162500, 15.487500, 3.441667, 5.162500, 5.162500)

# alpha_prop2 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.75*0.3, 0.25*0.3), # 0.225, 0.075
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop2 = c(5.162500, 15.487500, 3.441667, 7.743750, 2.581250)

# alpha_prop3 = dampack::dirichlet_params(
#   p.mean = c(0.15, 0.45 , 0.1, 0.9*0.3, 0.1*0.3), # 0.27, 0.03
#   sigma = c(0.06, 0.06, 0.03, 0.06, 0.06)
# )
# cannot install dampack in a reproductible environment (and not that necessary), so here the results of upper lines
alpha_prop3 = c(5.162500, 15.487500, 3.441667, 9.292500, 1.032500)


n = floor(nb_samples/3)
proportion = t(rbind(
  gtools::rdirichlet(n = n + (nb_samples - 3*n), alpha = alpha_prop1),
  gtools::rdirichlet(n = n, alpha = alpha_prop2),
  gtools::rdirichlet(n = n, alpha = alpha_prop3)
))

rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!

rm(alpha_prop1, alpha_prop2, alpha_prop3, n)
```

```{r, results="verbatim", eval=FALSE}
set.seed(10)

simu_proportion_pancreas = function(n, alpha = c(1.5, 4.5, 1, 3)){
  tmp_mix_Lvar = t(gtools::rdirichlet(n = n, alpha = alpha))
  tumor_rep = cbind(t(gtools::rdirichlet(n = n/3, alpha = c(0.5, 0.5))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.75, 0.25))),
              t(gtools::rdirichlet(n = n/3, alpha = c(0.9, 0.1))))
  mix_Lvar = matrix(nrow = 5, ncol = n)
  mix_Lvar [1:3, ] = tmp_mix_Lvar[1:3, ]
  mix_Lvar[4, ] = tumor_rep[1, ]*tmp_mix_Lvar[4, ]
  mix_Lvar[5, ] = tumor_rep[2, ]*tmp_mix_Lvar[4, ]

  return(mix_Lvar)
}

nb_samples = 60

proportion = simu_proportion_pancreas(n = nb_samples) # n -> number of samples
rownames(proportion) = c("endo", "fibro", "immune", "classic", "basal") # /!\ endo proportion unrealistic, modify!
```

## Ground truth - split {-}

```{r}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(proportion), size=round(ncol(proportion))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(proportion[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(proportion[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```

  
# Simulation of DNAm and RNA mixes from convolution Mix_met Mix_RNA

The following function adds noise to the simulation to make is harder to deconvolve, more life like.

In order to harder the deconvolution, here the added noise is randomly chosen from distribution with different parameters for different randomly selected subset of genes/met probes. 

Laplace or chi2 law can be used. Issue of chi2 is that it only add positive values. 


## Simulation Mix_RNA

Below code to generate only 1 simulation 


I start by generating the noise. I want correlation between genes so I simulate a correlated dataset (with the normal law family) with a correlation structure trying to be like the one of RNA in vitro data. It's the first version so it's very handcrafted


```{r generate_corr_data_rna, results="verbatim"}
div = 7.5 * nb_samples/30
a = 1:nb_samples / div

# nrow(rna_seq_src)/4 # If nrow(rna_seq_src) isn't a multiple of 4, it will not work

corr_dumb_data_rna = cbind(
  replicate(nrow(rna_seq_src)/8, rnorm(nb_samples)), # no corr
  replicate(nrow(rna_seq_src)*5/16, rnorm(nb_samples, mean = a/2)), # some slight more pos corr
  replicate(nrow(rna_seq_src)*5/16, rnorm(nb_samples, mean = -a/2)),  # some slight more neg corr
  replicate(nrow(rna_seq_src)/8, rnorm(nb_samples, mean = a)), # some slight more pos corr
  replicate(nrow(rna_seq_src)/8, rnorm(nb_samples, mean = -a))  # some slight more neg corr
)
# tmptmp = cor(corr_dumb_data_rna)
# tmptmp = tmptmp[lower.tri(tmptmp)]
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of two by two correlations between 10000 normally generated vector with mean a",
#   sub = paste0("where a is 1:", nb_samples, " / ", div),
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

```

Then we center and scale (by row) the data saw they keep the correlation structure but look like generated by a N(0,1) law. next step is to take they value by the C.D.F. so we have a uniform-like values.


```{r center_scale_pnorm_rna, results="verbatim"}
corr_dumb_data_rna = t(corr_dumb_data_rna)
mean_corr_dumb_data_rna = rowMeans(corr_dumb_data_rna)
c_corr_dumb_data_rna = corr_dumb_data_rna - tcrossprod(mean_corr_dumb_data_rna, rep(1, ncol(corr_dumb_data_rna)))  # centré
sd_corr_dumb_data_rna = sqrt(rowMeans(c_corr_dumb_data_rna^2))
sc_corr_dumb_data_rna = c_corr_dumb_data_rna/tcrossprod(sd_corr_dumb_data_rna, rep(1, ncol(corr_dumb_data_rna)))


unif_sc_corr_dumb_data_rna = pnorm(sc_corr_dumb_data_rna)
# tmptmp = cor(t(unif_sc_corr_dumb_data_rna))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of two by two correlations between 10000 unif transformed vector",
#   sub = paste0("where a is 1:30 / ", div),
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# # xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))
```

Now, unif-like values can be considered as p-value, we just have to define a law for each gene and the correlation structure is still conserved. Here for RNA data, I take a $\chi^2$ law center by its df and with 4 groups with different df.

```{r transform_to_gene_specific_law, results="verbatim"}
stopifnot(rownames(proportion) == colnames(rna_seq_src))

Mix_RNA =  as.matrix(rna_seq_src) %*% proportion #convolution of source matrix with proportion matrix
# subset_genes = subset_features(Mix_RNA, 4)

noise_rna = unif_sc_corr_dumb_data_rna
rownames(noise_rna) = rownames(Mix_RNA)

# ddl = 1
# noise_rna[subset_genes[[1]],] = apply(noise_rna[subset_genes[[1]],], 2, \(x){
#   qchisq(x, df = ddl) - ddl
# })
#   
# ddl = 2
# noise_rna[subset_genes[[2]],] = apply(noise_rna[subset_genes[[2]],], 2, \(x){
#   qchisq(x, df = ddl) - ddl
# })
# 
# ddl = 5
# noise_rna[subset_genes[[3]],] = apply(noise_rna[subset_genes[[3]],], 2, \(x){
#   qchisq(x, df = ddl) - ddl
# })
# 
# ddl = 10
# noise_rna[subset_genes[[4]],] = apply(noise_rna[subset_genes[[4]],], 2, \(x){
#   qchisq(x, df = ddl) - ddl
# })


# mean_epsilon_in_vitro_rna = readRDS(file = "01-add/simu_data_driven/mean_epsilon_in_vitro_rna_21000.rds") # decentrality
sd_epsilon_in_vitro_rna = readRDS(file = "01-additional/simu_data_driven/sd_epsilon_in_vitro_rna_21000.rds")

sd_epsilon_in_vitro_rna[sd_epsilon_in_vitro_rna > 3000] = 3000
stopifnot(identical(
  names(sd_epsilon_in_vitro_rna),
  rownames(noise_rna)
))

# decentrality
# For bias over 1% from a normal distrib with the sd of the specific gene, I take 2.5*sd_epsilon_in_vitro_rna with the same sign
# bias_exceeding = abs(mean_epsilon_in_vitro_rna) > 2.5*sd_epsilon_in_vitro_rna
# mean_epsilon_in_vitro_rna[bias_exceeding] = 2.5*sd_epsilon_in_vitro_rna[bias_exceeding]*sign(mean_epsilon_in_vitro_rna[bias_exceeding]) 

for (gene_i in 1:nrow(noise_rna)) {
  ddl_gene_i = sd_epsilon_in_vitro_rna[gene_i]
  noise_rna[gene_i,] = qchisq(noise_rna[gene_i,], df = ddl_gene_i) - ddl_gene_i # + mean_epsilon_in_vitro_rna[gene_i]
}

if (any(is.na(noise_rna))) {
  print(paste(sum(is.na(noise_rna)), "NaN generated by qchisq"))
  ddl_gene_i = 1
  noise_rna[is.na(noise_rna)] = qchisq(unif_sc_corr_dumb_data_rna[is.na(noise_rna)], df = ddl_gene_i) - ddl_gene_i
  stopifnot(sum(is.na(noise_rna)) == 0)
}


Mix_RNA = Mix_RNA + noise_rna

which_gene_neg = apply(Mix_RNA, 1, \(gene_i){any(gene_i<0)})
which_gene_neg = names(which_gene_neg[which(which_gene_neg)])

for (gene_i in which_gene_neg) {
  # ddl_gene_i = sd_epsilon_in_vitro_rna[gene_i]
  Mix_RNA[gene_i,] = Mix_RNA[gene_i,] - min(Mix_RNA[gene_i,]) + runif(nb_samples, min = 0, max = 1e-3)
}
# Mix_RNA[Mix_RNA<0] = runif(sum(Mix_RNA<0), min = 0, max = 5)

# tmp = Mix_RNA - as.matrix(rna_seq_src) %*% proportion
# tmptmp = cor(t(tmp))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of two by two correlations between 10000 unif transformed vector",
#   sub = paste0("where a is 1:30 / ", div),
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))
# 
# ggplot() +
#     geom_density(aes(x = rowMeans(tmp)))
# ggplot() +
#     geom_density(aes(x = rowMeans(tmp))) +
#     scale_x_continuous(limits = c(-5,5))
# ggplot() +
#     geom_density(aes(x = apply(tmp, 1, sd)))


rm(div, a, corr_dumb_data_rna, mean_corr_dumb_data_rna, c_corr_dumb_data_rna, sd_corr_dumb_data_rna, sc_corr_dumb_data_rna, unif_sc_corr_dumb_data_rna, noise_rna, ddl_gene_i, sd_epsilon_in_vitro_rna)#, bias_exceeding, mean_epsilon_in_vitro_rna)
```



## Simulation Mix_met

```{r generate_corr_data_met, results="verbatim"}
div = 7.5 * nb_samples/30
a = 1:nb_samples / div

# nrow(rna_seq_src)/4 # If nrow(rna_seq_src) isn't a multiple of 4, it will not work

# limite on pourrait determiner qu'elle mixture de loi normale permet de faire la densité de corr des vrai données

corr_dumb_data_met = cbind( 
  replicate(floor(nrow(dna_met_src)/16), rnorm(nb_samples, mean = -a/3)), # few neg corr       500
  replicate(ceiling(nrow(dna_met_src)*3/32), rnorm(nb_samples)), # no corr                     750 
  replicate(ceiling(nrow(dna_met_src)/8), rnorm(nb_samples, mean = a/3)), # few pos corr       1000
  replicate(floor(nrow(dna_met_src)*7/32), rnorm(nb_samples, mean = a/2)), # light pos corr    1750
  replicate(nrow(dna_met_src)/2, rnorm(nb_samples, mean = a)) # some slight more pos corr      4000
)
# tmptmp = cor(corr_dumb_data_met[,sample(1:nrow(dna_met_src), 20000)])
# tmptmp = tmptmp[lower.tri(tmptmp)]
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of two by two correlations between 20000 normally generated vector with mean a",
#   sub = paste0("where a is 1:", nb_samples, " / ", div),
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))

```



Then we center and scale (by row) the data saw they keep the correlation structure but look like generated by a N(0,1) law. next step is to take they value by the C.D.F. so we have a uniform-like values.


```{r center_scale_pnorm_met, results="verbatim"}
corr_dumb_data_met = t(corr_dumb_data_met)
mean_corr_dumb_data_met = rowMeans(corr_dumb_data_met)
c_corr_dumb_data_met = corr_dumb_data_met - tcrossprod(mean_corr_dumb_data_met, rep(1, ncol(corr_dumb_data_met)))  # centré
sd_corr_dumb_data_met = sqrt(rowMeans(c_corr_dumb_data_met^2))
sc_corr_dumb_data_met = c_corr_dumb_data_met/tcrossprod(sd_corr_dumb_data_met, rep(1, ncol(corr_dumb_data_met)))


unif_sc_corr_dumb_data_met = pnorm(sc_corr_dumb_data_met)
# tmptmp = cor(t(unif_sc_corr_dumb_data_met[sample(1:nrow(dna_met_src), 20000), ]))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of two by two correlations between 20000 unif transformed vector",
#   sub = paste0("where a is 1:30 / ", div),
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# # xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))
```


Now, unif-like values can be considered as p-value, we just have to define a law for each gene and the correlation structure is still conserved. Here for Methylation data, I take a Laplace law center to 0, with very law standard variations and again different groups with different sd. 

```{r transform_to_probe_specific_law, results="verbatim"}
stopifnot(rownames(proportion) == colnames(dna_met_src))

Mix_met = as.matrix(dna_met_src) %*% proportion #convolution of source matrix with proportion matrix
# subset_probes = subset_features(Mix_met, 4)

noise_met = unif_sc_corr_dumb_data_met
rownames(noise_met) = rownames(Mix_met)

# sigma = 1/15
# noise_met[subset_probes[[1]],] = apply(noise_met[subset_probes[[1]],], 2, \(x){
#   extraDistr::qlaplace(x, mu = 0, sigma = sigma)
# })
#   
# sigma = 1/20
# noise_met[subset_probes[[2]],] = apply(noise_met[subset_probes[[2]],], 2, \(x){
#   extraDistr::qlaplace(x, mu = 0, sigma = sigma)
# })
# 
# sigma = 1/30
# noise_met[subset_probes[[3]],] = apply(noise_met[subset_probes[[3]],], 2, \(x){
#   extraDistr::qlaplace(x, mu = 0, sigma = sigma)
# })
# 
# sigma = 1/50
# noise_met[subset_probes[[4]],] = apply(noise_met[subset_probes[[4]],], 2, \(x){
#   extraDistr::qlaplace(x, mu = 0, sigma = sigma)
# })


# mean_epsilon_in_vitro_met = readRDS(file = "01-add/simu_data_driven/mean_epsilon_in_vitro_met_820k.rds") # decentrality
sd_epsilon_in_vitro_met = readRDS(file = "01-additional/simu_data_driven/sd_epsilon_in_vitro_met_820k.rds")

stopifnot(identical(
  names(sd_epsilon_in_vitro_met),
  rownames(noise_met)
)) 

# decentrality
# For bias over 1% from a normal distrib with the sd of the specific gene, I take 2.5*sd_epsilon_in_vitro_rna with the same sign
# bias_exceeding = abs(mean_epsilon_in_vitro_met) > 2.5*sd_epsilon_in_vitro_met
# mean_epsilon_in_vitro_met[bias_exceeding] = 2.5*sd_epsilon_in_vitro_met[bias_exceeding]*sign(mean_epsilon_in_vitro_met[bias_exceeding])


for (probe_i in 1:nrow(noise_met)) {
  sigma_probe_i = sd_epsilon_in_vitro_met[probe_i]
  noise_met[probe_i,] = extraDistr::qlaplace(noise_met[probe_i,], mu = 0, sigma = sigma_probe_i) # + mean_epsilon_in_vitro_met[probe_i]
}

if (any(is.na(noise_met))) {
  print(paste(sum(is.na(noise_met)), "NaN generated by qlaplace"))
  sigma_probe_i = 1/20
  noise_met[is.na(noise_met)] = extraDistr::qlaplace(unif_sc_corr_dumb_data_met[is.na(noise_met)], mu = 0, sigma = sigma_probe_i)
  stopifnot(sum(is.na(noise_met)) == 0)
}


Mix_met = Mix_met + noise_met
Mix_met[Mix_met<0] = runif(sum(Mix_met<0), min = 0, max = 1e-03)
Mix_met[Mix_met>1] = runif(sum(Mix_met>1), min = 1-1e-03, max = 1)

# tmp_sample = sample(1:nrow(dna_met_src), 20000)
# tmp = Mix_met[tmp_sample,] - as.matrix(dna_met_src)[tmp_sample,] %*% proportion
# tmptmp = cor(t(tmp))
# tmptmp = tmptmp[lower.tri(tmptmp)]
# hist(
#   tmptmp,
#   nclass = 100,
#   main = "Distribution of two by two correlations between 10000 unif transformed vector",
#   sub = paste0("where a is 1:30 / ", div),
#   col = "tan",
#   ylab = "Density",
#   xlab = "Residual correlations",
#   freq = FALSE,
#   ylim = c(0, 3),
#   xlim = c(-1, 1),
#   bty = "l"
# )
# xx = seq(-1,1,0.01)
# lines(xx, dnorm(xx, sd = 1/sqrt(nb_samples)))
# 
# ggplot() +
#     geom_density(aes(x = rowMeans(tmp)))
# ggplot() +
#     geom_density(aes(x = apply(tmp, 1, sd)))


rm(div, a, corr_dumb_data_met, mean_corr_dumb_data_met, c_corr_dumb_data_met, sd_corr_dumb_data_met, sc_corr_dumb_data_met, unif_sc_corr_dumb_data_met, noise_met, sigma_probe_i, sd_epsilon_in_vitro_met, bias_exceeding, mean_epsilon_in_vitro_met)
```



## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=Mix_RNA[,idx_phaseUsers],
                              mix_met=Mix_met[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=Mix_RNA[,-idx_phaseUsers],
                             mix_met=Mix_met[,-idx_phaseUsers])
```

# Save into list

```{r, results="verbatim"}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```

