---
title: "HADACA3 - Generate invitrodata for data challenge hadaca"
author: "Elise Amblard, Florent Chuffart"
date: "`r Sys.Date()`"
output:
  rmarkdown::html_document:
---

This script generates in vitro data that can be used by the participants and in vitro data that will be used only during the scoring phase

```{r, include=(FALSE)}
knitr::opts_chunk$set(collapse=TRUE, comment="#>", fig.width=9, fig.height=6, eval=TRUE, echo=TRUE, results="verbatim")
options(knitr.duplicate.label = "allow")
```

# Parameters

Here we set up inputs and outputs files for the vignette.

```{r params}
if (!exists("source_file"))   {reference_file = "source_pdac.rds"} 
if (!exists("mixes_files"))       {mixes_files = c("mixes1_invitro_pdac.rds","mixes2_invitro_pdac.rds")} 
if (!exists("groundtruth_files")) {groundtruth_files = c("groundtruth1_invitro_pdac.rds","groundtruth2_invitro_pdac.rds")} 
```

# Objectives

This script generates:

- `r mixes_files` (a list with expression/methylation matrices `mix_rna` and `mix_met`), 
- `r groundtruth_files` (proportion matrix)

It takes as input reference data obtained from script 01.1_src_pdac.Rmd.

# Read reference data 

```{r reference}
ref = readRDS(paste0("../",source_file))
rna_seq_ref = ref$src_bulkRNA
dna_met_ref = ref$src_met
rm(ref)
gc()
```

# Ground truth and mixes matrices - read and split

## Ground truth - read {-}

```{r}
ground_truth = readRDS("~/projects/datashare/cometh_lot1/test_solution.rds")[[1]][,1:30]

colnames(rna_seq_ref)
rownames(ground_truth)
mask_ref = c("immune","basal","classic","immune","immune","endo","fibro","immune","immune")

ground_truth_with_mask = t(sapply(colnames(rna_seq_ref), function(celltype)
  if (sum(mask_ref==celltype)==1) {
    ground_truth[mask_ref==celltype,]
  } else {colSums(ground_truth[mask_ref==celltype,])}
))
```

## Ground truth - split {-}

```{r}
set.seed(20240)

idx_phaseUsers = sample(1:ncol(ground_truth_with_mask), size=round(ncol(ground_truth_with_mask))/2)

# idx_phaseUsers catches the indexes of the in vitro samples that will be visible for the challenge participants

saveRDS(ground_truth_with_mask[,idx_phaseUsers], paste0("../",groundtruth_files[1]))
saveRDS(ground_truth_with_mask[,-idx_phaseUsers], paste0("../",groundtruth_files[2]))
```

## Mixes matrices - read {-}

```{r}
# Loading the DESeq normalised RNA matrix
rna_measure = read.table("~/projects/datashare/cometh_lot1/transcriptome/gene_counts_normalised.txt")
rna_measure = as.matrix(rna_measure)
# Loading the filtered-normalised DNAm matrix
met_measure = readRDS("~/projects/datashare/cometh_lot1/methylation/test_data_met_new.rds")

rna_measure = rna_measure[,colnames(ground_truth_with_mask)]
met_measure = met_measure[,colnames(ground_truth_with_mask)]
```

## Mixes matrices - split {-}

```{r}
mixes_data_phaseUsers <- list(mix_rna=rna_measure[,idx_phaseUsers],
                              mix_met=met_measure[,idx_phaseUsers])
mixes_data_phaseEval <- list(mix_rna=rna_measure[,-idx_phaseUsers],
                             mix_met=met_measure[,-idx_phaseUsers])
```

# Save 

```{r}
saveRDS(mixes_data_phaseUsers, paste0("../",mixes_files[1]))
saveRDS(mixes_data_phaseEval, paste0("../",mixes_files[2]))
```


# Session Information

```{r, results="verbatim"}
sessionInfo()
```
