---
title: "HADACA3 Framework - Supplementary figure 6, 8, 9 and 10"
author: ""
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---



```{r, echo=FALSE, eval=TRUE}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=12, fig.height=12, eval=TRUE, echo=TRUE, results="verbatim", dpi=75)
```

```{r re_loading_pckgs, echo=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
theme_set(theme_light())
library(wesanderson)
library(RColorBrewer)
library(ComplexHeatmap)
```

# Load data

```{r load_table}

weights_corrected = c(rep(1/4*1/2,2),
                    rep(1/4*1/4,4),
                    rep(1/4*1/4,4),
                    rep(1/4*1/2,2))

metric_cols = c("pearson_row_norm","spearman_row_norm",
                "pearson_tot_norm","pearson_col_norm","spearman_tot_norm","spearman_col_norm",
                "rmse_norm","mae_norm","aitchison_norm","jsd_norm",
                "sdid_norm","aid_norm")

weighgeomMean = function(x) {
  prod(mapply(function(a,b) a^b, a=x, b=weights_corrected), na.rm=T)^(1/sum(weights_corrected))
}


if(!exists("results_li")){
  print("reading results li csv")
  
  all_functions_li = c(
    'preprocessing_mixRNA', 'feature_selection_mixRNA',
    'preprocessing_RNA', 'feature_selection_RNA',
    'preprocessing_scRNA', 'feature_selection_scRNA', 'deconvolution_rna',
    'preprocessing_mixMET', 'feature_selection_mixMET',
    'preprocessing_MET', 'feature_selection_MET', 'deconvolution_met',
    'late_integration'
  )
  results_li_all_dataset = rbind(
    read.csv(file = gzfile("results_li_invitro_invivo_insilicopsuedobulk.csv.gz")),
    read.csv(file = gzfile("results_li_EMFA_Copule_Immuneprop1.csv.gz")),
    read.csv(file = gzfile("results_li_NODEP6_NODEP4_NODEP1.csv.gz"))
  ) 
  # results_li_all_dataset = read.csv(file = gzfile("../old_results_li_ei/results_li.csv.gz"))
 # results_li_all_dataset = results_li_all_dataset[-(results_li_all_dataset$dataset == "51847637672785"),] % MAG:incorrect
  results_li_all_dataset <- results_li_all_dataset[
  results_li_all_dataset$dataset != "51847637672785", ]
  results_li_all_dataset = results_li_all_dataset[!is.na(results_li_all_dataset$dataset),]
  results_li_all_dataset = results_li_all_dataset[!is.na(results_li_all_dataset$feature_selection_mixRNA),]
  
  
  results_li_all_dataset[all_functions_li] = lapply(results_li_all_dataset[all_functions_li], as.factor)
  
  results_li = results_li_all_dataset %>% 
    filter(feature_selection_mixMET != "Toastnbfs") %>% 
    group_by(
      preprocessing_mixRNA, feature_selection_mixRNA,
      preprocessing_RNA, feature_selection_RNA,
      preprocessing_scRNA, feature_selection_scRNA, deconvolution_rna,
      preprocessing_mixMET, feature_selection_mixMET,
      preprocessing_MET, feature_selection_MET, deconvolution_met,
      late_integration
    ) %>% 
    summarise(
      across(where(is.numeric), \(x) mean(x, na.rm = TRUE)), .groups = "drop"
    ) %>% 
    as.data.frame()

  
  results_li = results_li[!is.na(results_li$feature_selection_RNA),] # One NA
  
  results_li$score_aggreg= apply(   #_corrected 
    results_li[, metric_cols],
    1,weighgeomMean
  )

}

if(!exists("results_ei")){
  print("reading reulsts ei csv")
  
  all_functions_ei <- c(
    'preprocessing_mixRNA', 'feature_selection_mixRNA',
    'preprocessing_RNA', 'feature_selection_RNA',
    'preprocessing_scRNA', 'feature_selection_scRNA',
    'preprocessing_mixMET', 'feature_selection_mixMET',
    'preprocessing_MET', 'feature_selection_MET',
    'early_integration', 'deconvolution'
  )
  
  results_ei_all_dataset = rbind(
    read.csv(file = gzfile("results_ei_invitro_invivo_insilicopsuedobulk.csv.gz")),
    read.csv(file = gzfile("results_ei_EMFA_Copule_Immuneprop1.csv.gz")),
    read.csv(file = gzfile("results_ei_NODEP6_NODEP4_NODEP1.csv.gz"))
  )  
  # results_ei_all_dataset = read.csv(file = gzfile("../old_results_li_ei/merged_results_ei.csv.gz"))
  
  tmp = results_ei_all_dataset[results_ei_all_dataset$score_aggreg == 0,]
  apply(tmp[,c("dataset", "ref", all_functions_ei)], 2, \(col_i){unique(col_i)})
  # On a toutes les combinaisons possibles et tout les datasets, donc ce n'est pas liée à un truc en particulier 
  # J'ai regardée le fichier 06, et on ne véréfie pas la somme à 1 mais on force bien à ce que tout les min soit à 1e-9
  
  
  results_ei_all_dataset[all_functions_ei] <- lapply(results_ei_all_dataset[all_functions_ei], as.factor)

  results_ei = results_ei_all_dataset %>% 
    filter(feature_selection_mixMET != "Toastnbfs") %>% 
    group_by(
      preprocessing_mixRNA, feature_selection_mixRNA,
      preprocessing_RNA, feature_selection_RNA,
      preprocessing_scRNA, feature_selection_scRNA,
      preprocessing_mixMET, feature_selection_mixMET,
      preprocessing_MET, feature_selection_MET,
      deconvolution, early_integration
    ) %>% 
    summarise(
      across(where(is.numeric), \(x) mean(x, na.rm = TRUE)), .groups = "drop"
    ) %>% 
    as.data.frame()
  
  
  results_ei = results_ei[results_ei$score_aggreg != 0,]





  # recompute corrected aggregated score
  results_ei$score_aggreg= apply(   #_corrected 
    results_ei[, metric_cols],
    1,weighgeomMean
  )

}
```


```{r ei_levels_def}
levels(factor(results_ei$preprocessing_RNA))
levels(factor(results_ei$feature_selection_RNA))

levels(factor(results_ei$preprocessing_scRNA))
levels(factor(results_ei$feature_selection_scRNA))

levels(factor(results_ei$preprocessing_MET))
levels(factor(results_ei$feature_selection_MET))

levels(factor(results_ei$early_integration))

levels(factor(results_ei$deconvolution))


results_ei$preprocessing_RNA = factor(results_ei$preprocessing_RNA, levels = c("ppID", "LogNorm", "Scale"))
results_ei$feature_selection_RNA = factor(results_ei$feature_selection_RNA, levels = c("fsID", "SCcluster", "scpseudobulk", "Toastbulknbfs", "Toastvst"))
results_ei$preprocessing_scRNA = factor(results_ei$preprocessing_scRNA, levels = c("ppID", "sccluster", "scConcat", "LogNorm", "Scale", "scbinarypseudobulklog"))
results_ei$feature_selection_scRNA = factor(results_ei$feature_selection_scRNA, levels = c("fsID", "SCcluster", "scpseudobulk", "Toastbulknbfs", "Toastvst"))
results_ei$preprocessing_MET = factor(results_ei$preprocessing_MET, levels = c("ppID", "LogNorm", "Scale"))
results_ei$feature_selection_MET = factor(results_ei$feature_selection_MET, levels = c("fsID", "maxdiscriminant", "mostmethylated", "SPLSDA", "Toastpercent")) # , "Toastnbfs"
results_ei$early_integration = factor(results_ei$early_integration, levels = c("concatnoscale", "concatscale", "Kernel", "omicade4bulk", "OT"))
results_ei$deconvolution = factor(results_ei$deconvolution, levels = c("nnls", "lm", "nnlslargeref", "RLR", "RLRnnls", "RLRpoisson", "epic"))
```

```{r li_levels_def}
levels(factor(results_li$preprocessing_RNA))
levels(factor(results_li$feature_selection_RNA))

levels(factor(results_li$preprocessing_scRNA))
levels(factor(results_li$feature_selection_scRNA))

levels(factor(results_li$preprocessing_MET))
levels(factor(results_li$feature_selection_MET))

levels(factor(results_li$deconvolution_rna))
levels(factor(results_li$deconvolution_met))

levels(factor(results_li$late_integration))


results_li$preprocessing_RNA = factor(results_li$preprocessing_RNA, levels = c("ppID", "LogNorm", "Scale", "nopp")) # , "CPM"
results_li$feature_selection_RNA = factor(results_li$feature_selection_RNA, levels = c("fsID", "SCcluster",  "scpseudobulk", "Toastbulknbfs", "Toastvst", "nofs")) # , "scSPLSDA"
results_li$preprocessing_scRNA = factor(results_li$preprocessing_scRNA, levels = c("ppID", "LogNorm", "Scale", "scbinarypseudobulklog", "sccluster", "scConcat", "nopp")) # , "CPM", "scCCAintegration", "scScale"
results_li$feature_selection_scRNA = factor(results_li$feature_selection_scRNA, levels = c("fsID", "SCcluster",  "scpseudobulk", "Toastbulknbfs", "Toastvst", "nofs")) # , "scSPLSDA"
results_li$preprocessing_MET = factor(results_li$preprocessing_MET, levels = c("ppID", "LogNorm", "Scale", "nopp")) # , "normalize"
results_li$feature_selection_MET = factor(results_li$feature_selection_MET, levels = c("fsID", "maxdiscriminant", "mostmethylated", "SPLSDA", "Toastpercent", "nofs")) # , "Toastnbfs"
results_li$deconvolution_rna = factor(results_li$deconvolution_rna, levels = c("nnls", "lm",  "nnlslargeref", "RLR", "RLRnnls", "RLRpoisson", "epic", "node"))
results_li$deconvolution_met = factor(results_li$deconvolution_met, levels = c("nnls", "lm",  "nnlslargeref", "RLR", "RLRnnls", "RLRpoisson", "epic", "node"))
results_li$late_integration = factor(results_li$late_integration, levels = c("limean", "TunedJ", "limeanRMSE", "OnlyMet", "OnlyRna")) # , "liCtSens"
```


```{r color_choices}

# wes_palette("Royal2") # for modules
module_pal = wes_palette("Royal2")[c(2, 3, 5, 4, 1)]
names(module_pal) = c("preprocessing", "feature_selection", "deconvolution", "early_integration", "late_integration")




# generated with Claude

all_levels <- unique(c(
  levels(results_ei$preprocessing_RNA),
  levels(results_ei$feature_selection_RNA),
  levels(results_ei$preprocessing_scRNA),
  levels(results_ei$feature_selection_scRNA),
  levels(results_ei$preprocessing_MET),
  levels(results_ei$feature_selection_MET),
  levels(results_ei$early_integration),
  levels(results_ei$deconvolution),
  levels(results_li$preprocessing_RNA),
  levels(results_li$feature_selection_RNA),
  levels(results_li$preprocessing_scRNA),
  levels(results_li$feature_selection_scRNA),
  levels(results_li$preprocessing_MET),
  levels(results_li$feature_selection_MET),
  levels(results_li$deconvolution_rna),
  levels(results_li$deconvolution_met),
  levels(results_li$late_integration)
))

id_levels    <- all_levels[grepl(c("ID"), all_levels) | grepl(c("nopp"), all_levels) | grepl(c("nofs"), all_levels) | grepl(c("node"), all_levels) | grepl(c("OnlyRna"), all_levels) | grepl(c("OnlyMet"), all_levels)]
other_levels <- all_levels[!(grepl(c("ID"), all_levels) | grepl(c("nopp"), all_levels) | grepl(c("nofs"), all_levels) | grepl(c("node"), all_levels))]
n_other <- length(other_levels)

palette_colors <- colorRampPalette(
  c(brewer.pal(7, "Set2"),
    brewer.pal(7, "Dark2"),
    brewer.pal(8, "Set3"))
)(n_other)


level_colors <- c(
  setNames(rep("grey60", length(id_levels)), id_levels),
  setNames(palette_colors, other_levels)
)

# # Verification
# scales::show_col(level_colors, ncol = 6)
# print(level_colors)
```



# Supp Figure 6: Correlation between metrics


```{r cor_plot, fig.height=9}
# results_li[results_li=="NaN"] = NA
row_rm = apply(results_li,1,function(x) any(is.na(x)))
results_li_rm = results_li[!row_rm,-(1:(length(all_functions_li)))]

# results_ei[results_ei=="NaN"] = NA
row_rm = apply(results_ei,1,function(x) any(is.na(x)))
results_ei_rm = results_ei[!row_rm,-(1:(length(all_functions_ei)))]


score_aggreg_order = grep("score_aggreg", colnames(results_ei_rm))
normalised_metrics = grep("norm", colnames(results_ei_rm))
unnormalised_metrics = (1:ncol(results_ei_rm))[-c(score_aggreg_order, normalised_metrics)]
corr_metric = c(grep("pearson", colnames(results_ei_rm)), grep("spearman", colnames(results_ei_rm)))

correlations = cor(rbind(
  results_li_rm[,c(
    score_aggreg_order,
    normalised_metrics[!(normalised_metrics %in% corr_metric)],
    normalised_metrics[normalised_metrics %in% corr_metric]
    # unnormalised_metrics[!(unnormalised_metrics %in% corr_metric)],
    # unnormalised_metrics[unnormalised_metrics %in% corr_metric]
  )],
  results_ei_rm[,c(
    score_aggreg_order,
    normalised_metrics[!(normalised_metrics %in% corr_metric)],
    normalised_metrics[normalised_metrics %in% corr_metric]
    # unnormalised_metrics[!(unnormalised_metrics %in% corr_metric)],
    # unnormalised_metrics[unnormalised_metrics %in% corr_metric]
  )]
))
corrplot::corrplot(correlations, method='ellipse', type='full', outline=T, tl.col='black')
```


# Supp Figure 8: Preprocessing

## Violin plots

```{r fun_arranged_violinplot}
arranged_violinplot = function(data, x_var, inte, score = "score_aggreg", module_palette = module_pal, level_palette = level_colors){
  
  if (x_var != "early_integration" && x_var != "late_integration") {
    data = data[data$preprocessing_mixRNA != "nopp",]
    data = data[data$feature_selection_mixRNA != "nofs",]
    data = data[data$preprocessing_RNA != "nopp",]
    data = data[data$feature_selection_RNA != "nofs",]
    data = data[data$preprocessing_scRNA != "nopp",]
    data = data[data$feature_selection_scRNA != "nofs",]
    data = data[data$preprocessing_mixMET != "nopp",]
    data = data[data$feature_selection_mixMET != "nofs",]
    data = data[data$preprocessing_MET != "nopp",]
    data = data[data$feature_selection_MET != "nofs",]
  }
  
  search_module = which(sapply(names(module_palette), \(module_i){grepl(module_i, x = x_var)}) == 1) 
  color_module = module_pal[search_module]
  
  
  order_level_x_var = sapply(levels(data[,x_var]), \(level_i){
    median(data[data[,x_var] == level_i, score], na.rm = TRUE)
  })
  # if we look at mae, we want to look at the minimum, otherwise (score aggreg and cor) the max
  order_level_x_var = order_level_x_var[order(order_level_x_var, decreasing = !(score == "mae"))]
  data[,x_var] = factor(data[,x_var], levels = names(order_level_x_var))
  
  #ADD MEDIAN 
   median_data <- data %>%
    group_by(!!sym(x_var)) %>%
    summarise(median_val = median(!!sym(score), na.rm = TRUE)) %>%
    rename(!!x_var := !!x_var)
   
  print(car::leveneTest(y = data[,score], data[,x_var]))
  
  
  aes_base = aes_string(x = x_var, y = score, fill = x_var)
  aes_base_median = aes_string(
    x = x_var, 
    y = sapply(levels(data[,x_var]), \(level_i){
      median(data[data[,x_var] == level_i, score])
    })
  )
  ggplot(data[,c(x_var, score)]) +
    # geom_boxplot(aes_base) +
    geom_violin(aes_base, draw_quantiles = 0.5) +
    # median point manually displayed
    stat_summary(aes_base, fun = median, geom = "point", shape = "-", size = 5, color = "black") + 
     geom_text(                                        
      data  = median_data,
      aes_string(x = x_var, y = "median_val",
                 label = "round(median_val, 2)"),
      vjust = -0.5,
      hjust = 0.5,
      size  = 6,
      color = "black"
    ) +
    ylim(0, 1) +
    labs(
      title = paste("Aggregated score for", x_var), #, "within early integration strategies"
      subtitle = paste("within", inte, "integration strategies"), # "ordered by median of aggregated score (red horizontal line)",
      x = element_blank(),
      y = "Aggregated score",
    ) +
    scale_fill_manual(
      values = level_palette[levels(data[,x_var])]
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      panel.border = element_rect(color = color_module, fill = NA, linewidth = 3),
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "none" # "bottom"
    )
  
}



all_functions_ei

pdf("violinplots_early_PP_RNA.pdf", width = 5, height = 5)
arranged_violinplot(data = results_ei, x_var = "preprocessing_RNA", inte = "early")
dev.off()

pdf("violinplots_early_PP_DNAm.pdf", width = 5, height = 5)
arranged_violinplot(data = results_ei, x_var = "preprocessing_MET", inte = "early")
dev.off()



all_functions_li

pdf("violinplots_late_PP_RNA.pdf", width = 5, height = 5)
arranged_violinplot(data = results_li, x_var = "preprocessing_RNA", inte = "late")
dev.off()

pdf("violinplots_late_PP_DNAm.pdf", width = 5, height = 5)
arranged_violinplot(data = results_li, x_var = "preprocessing_MET", inte = "late")
dev.off()
```


## Pairwise omics interaction for preprocessing

### Early_integration

```{r PP_pairwise_interaction_EI}
col_fun = circlize::colorRamp2(
  c(-0.075, -0.025, 0, 0.025, 0.075),
  # wes_palette("Zissou1")[1:5]
  # c("blue", "white", "red")
  RColorBrewer::brewer.pal(5, "RdBu")[5:1]
)


## Overall EI median score
overall_median_ei_score_aggreg = median(results_ei$score_aggreg)


results_ei %>% 
  group_by(preprocessing_RNA, preprocessing_MET) %>% 
  summarise(n = n(), median = median(score_aggreg)) %>% print(n = 25)

med_score_aggr_PP = matrix(
  data = results_ei %>% 
      group_by(preprocessing_RNA, preprocessing_MET) %>% 
      summarise(median = median(score_aggreg)) %>% 
      pull(median) - overall_median_ei_score_aggreg,
  byrow = TRUE, # Methylation is in columns
  nrow = length(levels(results_ei$preprocessing_RNA)),
  ncol = length(levels(results_ei$preprocessing_MET)),
  dimnames = list(
    levels(results_ei$preprocessing_RNA),
    levels(results_ei$preprocessing_MET)
  )
)

pdf("heatmap_PP_early.pdf", width = 3, height = 3)
Heatmap(
  med_score_aggr_PP,
  name = "median aggr score",
  # title = "Omcis preprocessing comparison on median aggregated score for early inte",
  col = col_fun,
  column_title = "Preprocessing DNAm",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_title = "Preprocessing RNA",
  row_title_gp = gpar(fontsize = 14, fontface = "bold"),
  # cell_fun = cell_fun_maker(mat_PP_rna_pvalue),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_heatmap_legend = TRUE,
  na_col = "grey90",
  # border = TRUE,
  rect_gp = gpar(col = "white", lwd = 2)
)
dev.off()
```


### Late_integration

```{r PP_pairwise_interaction_LI}
col_fun = circlize::colorRamp2(
  c(-0.075, -0.025, 0, 0.025, 0.075),
  # wes_palette("Zissou1")[1:5]
  # c("blue", "white", "red")
  RColorBrewer::brewer.pal(5, "RdBu")[5:1]
)


## Overall LI median score
overall_median_li_score_aggreg = median(results_li$score_aggreg, na.rm = TRUE)


results_li %>% 
  group_by(preprocessing_RNA, preprocessing_MET) %>% 
  summarise(n = n(), median = median(score_aggreg)) %>% print(n = 25)

med_score_aggr_PP = matrix(
  data = c(
    results_li %>% 
      group_by(preprocessing_RNA, preprocessing_MET) %>% 
      summarise(median = median(score_aggreg)) %>% 
      pull(median) - overall_median_li_score_aggreg,
    NA # Since there is no nopp/nopp combination 
  ),
  byrow = TRUE, # Methylation is in columns
  nrow = length(levels(results_li$preprocessing_RNA)),
  ncol = length(levels(results_li$preprocessing_MET)),
  dimnames = list(
    levels(results_li$preprocessing_RNA),
    levels(results_li$preprocessing_MET)
  )
)

pdf("heatmap_PP_late.pdf", width = 3, height = 3)
Heatmap(
  med_score_aggr_PP[-nrow(med_score_aggr_PP), -ncol(med_score_aggr_PP)],
  name = "median aggr score",
  col = col_fun,
  column_title = "Preprocessing DNAm",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_title = "Preprocessing RNA",
  row_title_gp = gpar(fontsize = 14, fontface = "bold"),
  # cell_fun = cell_fun_maker(mat_PP_rna_pvalue),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  # row_split = c(1,1,1,1,2),
  # column_split = c(1,1,1,1,2),
  show_heatmap_legend = TRUE,
  na_col = "grey90",
  # border = TRUE,
  #filename = "heatmap_best_pipeline.pdf",
  rect_gp = gpar(col = "white", lwd = 2)
)
dev.off()
```


# Supp Figure 9: Feature selection

## Violin plots

```{r fun_arranged_violinplot}
arranged_violinplot = function(data, x_var, inte, score = "score_aggreg", module_palette = module_pal, level_palette = level_colors){
  
  if (x_var != "early_integration" && x_var != "late_integration") {
    data = data[data$preprocessing_mixRNA != "nopp",]
    data = data[data$feature_selection_mixRNA != "nofs",]
    data = data[data$preprocessing_RNA != "nopp",]
    data = data[data$feature_selection_RNA != "nofs",]
    data = data[data$preprocessing_scRNA != "nopp",]
    data = data[data$feature_selection_scRNA != "nofs",]
    data = data[data$preprocessing_mixMET != "nopp",]
    data = data[data$feature_selection_mixMET != "nofs",]
    data = data[data$preprocessing_MET != "nopp",]
    data = data[data$feature_selection_MET != "nofs",]
  }
  
  search_module = which(sapply(names(module_palette), \(module_i){grepl(module_i, x = x_var)}) == 1) 
  color_module = module_pal[search_module]
  
  
  order_level_x_var = sapply(levels(data[,x_var]), \(level_i){
    median(data[data[,x_var] == level_i, score], na.rm = TRUE)
  })
  # if we look at mae, we want to look at the minimum, otherwise (score aggreg and cor) the max
  order_level_x_var = order_level_x_var[order(order_level_x_var, decreasing = !(score == "mae"))]
  data[,x_var] = factor(data[,x_var], levels = names(order_level_x_var))
  
  #ADD MEDIAN 
   median_data <- data %>%
    group_by(!!sym(x_var)) %>%
    summarise(median_val = median(!!sym(score), na.rm = TRUE)) %>%
    rename(!!x_var := !!x_var)
   
  print(car::leveneTest(y = data[,score], data[,x_var]))
  
  
  aes_base = aes_string(x = x_var, y = score, fill = x_var)
  aes_base_median = aes_string(
    x = x_var, 
    y = sapply(levels(data[,x_var]), \(level_i){
      median(data[data[,x_var] == level_i, score])
    })
  )
  ggplot(data[,c(x_var, score)]) +
    # geom_boxplot(aes_base) +
    geom_violin(aes_base, draw_quantiles = 0.5) +
    # median point manually displayed
    stat_summary(aes_base, fun = median, geom = "point", shape = "-", size = 5, color = "black") + 
     geom_text(                                        
      data  = median_data,
      aes_string(x = x_var, y = "median_val",
                 label = "round(median_val, 2)"),
      vjust = -0.5,
      hjust = 0.5,
      size  = 6,
      color = "black"
    ) +
    ylim(0, 1) +
    labs(
      title = paste("Aggregated score for", x_var), #, "within early integration strategies"
      subtitle = paste("within", inte, "integration strategies"), # "ordered by median of aggregated score (red horizontal line)",
      x = element_blank(),
      y = "Aggregated score",
    ) +
    scale_fill_manual(
      values = level_palette[levels(data[,x_var])]
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      panel.border = element_rect(color = color_module, fill = NA, linewidth = 3),
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "none" # "bottom"
    )
  
}



all_functions_ei

pdf("violinplots_early_FS_RNA.pdf", width = 5, height = 5)
arranged_violinplot(data = results_ei, x_var = "feature_selection_RNA", inte = "early")
dev.off()

pdf("violinplots_early_FS_DNAm.pdf", width = 5, height = 5)
arranged_violinplot(data = results_ei, x_var = "feature_selection_MET", inte = "early")
dev.off()



all_functions_li

pdf("violinplots_late_FS_RNA.pdf", width = 5, height = 5)
arranged_violinplot(data = results_li, x_var = "feature_selection_RNA", inte = "late")
dev.off()

pdf("violinplots_late_FS_DNAm.pdf", width = 5, height = 5)
arranged_violinplot(data = results_li, x_var = "feature_selection_MET", inte = "late")
dev.off()
```


## Pairwise omics interaction for preprocessing

### Early_integration

```{r PP_pairwise_interaction_EI}
col_fun = circlize::colorRamp2(
  c(-0.075, -0.025, 0, 0.025, 0.075),
  # wes_palette("Zissou1")[1:5]
  # c("blue", "white", "red")
  RColorBrewer::brewer.pal(5, "RdBu")[5:1]
)


## Overall EI median score
overall_median_ei_score_aggreg = median(results_ei$score_aggreg)


results_ei %>% 
  group_by(feature_selection_RNA, feature_selection_MET) %>% 
  summarise(n = n(), median = median(score_aggreg)) %>% print(n = 25)

med_score_aggr_FS = matrix(
  data = results_ei %>% 
      group_by(feature_selection_RNA, feature_selection_MET) %>% 
      summarise(median = median(score_aggreg)) %>% 
      pull(median) - overall_median_ei_score_aggreg,
  byrow = TRUE, # Methylation is in columns
  nrow = length(levels(results_ei$feature_selection_RNA)),
  ncol = length(levels(results_ei$feature_selection_MET)),
  dimnames = list(
    levels(results_ei$feature_selection_RNA),
    levels(results_ei$feature_selection_MET)
  )
)

pdf("heatmap_FS_early.pdf", width = 3, height = 3)
Heatmap(
  med_score_aggr_FS,
  name = "median aggr score",
  col = col_fun,
  column_title = "Feature selecion DNAm",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_title = "Feature selecion RNA",
  row_title_gp = gpar(fontsize = 14, fontface = "bold"),
  # cell_fun = cell_fun_maker(mat_PP_rna_pvalue),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  # row_split = c(1,1,1,2),
  # column_split = c(1,2),
  show_heatmap_legend = TRUE,
  na_col = "grey90",
  # border = TRUE,
  rect_gp = gpar(col = "white", lwd = 2)
)
dev.off()
```


### Late_integration

```{r PP_pairwise_interaction_LI}
col_fun = circlize::colorRamp2(
  c(-0.075, -0.025, 0, 0.025, 0.075),
  # wes_palette("Zissou1")[1:5]
  # c("blue", "white", "red")
  RColorBrewer::brewer.pal(5, "RdBu")[5:1]
)


## Overall LI median score
overall_median_li_score_aggreg = median(results_li$score_aggreg, na.rm = TRUE)


results_li %>% 
  group_by(feature_selection_RNA, feature_selection_MET) %>% 
  summarise(n = n(), median = median(score_aggreg)) %>% print(n = 25)

med_score_aggr_FS = matrix(
  data = c(
    results_li %>% 
      group_by(feature_selection_RNA, feature_selection_MET) %>% 
      summarise(median = median(score_aggreg)) %>% 
      pull(median) - overall_median_li_score_aggreg,
    NA # Since there is no nopp/nopp combination 
  ),
  byrow = TRUE, # Methylation is in columns
  nrow = length(levels(results_li$feature_selection_RNA)),
  ncol = length(levels(results_li$feature_selection_MET)),
  dimnames = list(
    levels(results_li$feature_selection_RNA),
    levels(results_li$feature_selection_MET)
  )
)

pdf("heatmap_FS_late.pdf", width = 3, height = 3)
Heatmap(
  med_score_aggr_FS[-nrow(med_score_aggr_FS), -ncol(med_score_aggr_FS)],
  name = "fsID",
  col = col_fun,
  column_title = "Feature selecion DNAm",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_title = "Feature selecion RNA",
  row_title_gp = gpar(fontsize = 14, fontface = "bold"),
  # cell_fun = cell_fun_maker(mat_PP_rna_pvalue),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  # row_split = c(1,1,1,2),
  # column_split = c(1,2),
  show_heatmap_legend = TRUE,
  na_col = "grey90",
  # border = TRUE,
  rect_gp = gpar(col = "white", lwd = 2)
)
dev.off()
```


# Supp Figure 10: Deconvolution

## Violin plots

```{r fun_arranged_violinplot}
arranged_violinplot = function(data, x_var, inte, score = "score_aggreg", module_palette = module_pal, level_palette = level_colors){
  
  if (x_var != "early_integration" && x_var != "late_integration") {
    data = data[data$preprocessing_mixRNA != "nopp",]
    data = data[data$feature_selection_mixRNA != "nofs",]
    data = data[data$preprocessing_RNA != "nopp",]
    data = data[data$feature_selection_RNA != "nofs",]
    data = data[data$preprocessing_scRNA != "nopp",]
    data = data[data$feature_selection_scRNA != "nofs",]
    data = data[data$preprocessing_mixMET != "nopp",]
    data = data[data$feature_selection_mixMET != "nofs",]
    data = data[data$preprocessing_MET != "nopp",]
    data = data[data$feature_selection_MET != "nofs",]
  }
  
  search_module = which(sapply(names(module_palette), \(module_i){grepl(module_i, x = x_var)}) == 1) 
  color_module = module_pal[search_module]
  
  
  order_level_x_var = sapply(levels(data[,x_var]), \(level_i){
    median(data[data[,x_var] == level_i, score], na.rm = TRUE)
  })
  # if we look at mae, we want to look at the minimum, otherwise (score aggreg and cor) the max
  order_level_x_var = order_level_x_var[order(order_level_x_var, decreasing = !(score == "mae"))]
  data[,x_var] = factor(data[,x_var], levels = names(order_level_x_var))
  
  #ADD MEDIAN 
   median_data <- data %>%
    group_by(!!sym(x_var)) %>%
    summarise(median_val = median(!!sym(score), na.rm = TRUE)) %>%
    rename(!!x_var := !!x_var)
   
  print(car::leveneTest(y = data[,score], data[,x_var]))
  
  
  aes_base = aes_string(x = x_var, y = score, fill = x_var)
  aes_base_median = aes_string(
    x = x_var, 
    y = sapply(levels(data[,x_var]), \(level_i){
      median(data[data[,x_var] == level_i, score])
    })
  )
  ggplot(data[,c(x_var, score)]) +
    # geom_boxplot(aes_base) +
    geom_violin(aes_base, draw_quantiles = 0.5) +
    # median point manually displayed
    stat_summary(aes_base, fun = median, geom = "point", shape = "-", size = 5, color = "black") + 
     geom_text(                                        
      data  = median_data,
      aes_string(x = x_var, y = "median_val",
                 label = "round(median_val, 2)"),
      vjust = -0.5,
      hjust = 0.5,
      size  = 6,
      color = "black"
    ) +
    ylim(0, 1) +
    labs(
      title = paste("Aggregated score for", x_var), #, "within early integration strategies"
      subtitle = paste("within", inte, "integration strategies"), # "ordered by median of aggregated score (red horizontal line)",
      x = element_blank(),
      y = "Aggregated score",
    ) +
    scale_fill_manual(
      values = level_palette[levels(data[,x_var])]
    ) +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 12),
      panel.border = element_rect(color = color_module, fill = NA, linewidth = 3),
      plot.title = element_text(hjust = 0.5),
      plot.subtitle = element_text(hjust = 0.5),
      legend.position = "none" # "bottom"
    )
  
}



all_functions_ei

pdf("violinplots_early_deconv.pdf", width = 5, height = 5)
arranged_violinplot(data = results_ei, x_var = "deconvolution", inte = "early")
dev.off()



all_functions_li

pdf("violinplots_late_deconv_RNA.pdf", width = 5, height = 5)
arranged_violinplot(data = results_li, x_var = "deconvolution_rna", inte = "late")
dev.off()

pdf("violinplots_late_deconv_DNAm.pdf", width = 5, height = 5)
arranged_violinplot(data = results_li, x_var = "deconvolution_met", inte = "late")
dev.off()
```


## Pairwise omics interaction for preprocessing

### Late_integration

```{r PP_pairwise_interaction_LI}
col_fun = circlize::colorRamp2(
  c(-0.075, -0.025, 0, 0.025, 0.075),
  # wes_palette("Zissou1")[1:5]
  # c("blue", "white", "red")
  RColorBrewer::brewer.pal(5, "RdBu")[5:1]
)


## Overall LI median score
overall_median_li_score_aggreg = median(results_li$score_aggreg, na.rm = TRUE)


results_li %>% 
  group_by(deconvolution_rna, deconvolution_met) %>% 
  summarise(n = n(), median = median(score_aggreg)) %>% print(n = 25)

med_score_aggr_deconv = matrix(
  data = c(
    results_li %>% 
      group_by(deconvolution_rna, deconvolution_met) %>% 
      summarise(median = median(score_aggreg)) %>% 
      pull(median) - overall_median_li_score_aggreg,
    NA # Since there is no nopp/nopp combination 
  ),
  byrow = TRUE, # Methylation is in columns
  nrow = length(levels(results_li$deconvolution_rna)),
  ncol = length(levels(results_li$deconvolution_met)),
  dimnames = list(
    levels(results_li$deconvolution_rna),
    levels(results_li$deconvolution_met)
  )
)

pdf("heatmap_deconv_late.pdf", width = 3, height = 3)
Heatmap(
  med_score_aggr_deconv[-nrow(med_score_aggr_deconv), -ncol(med_score_aggr_deconv)],
  name = "median aggr score",
  col = col_fun,
  column_title = "Deconvolution DNAm",
  column_title_gp = gpar(fontsize = 14, fontface = "bold"),
  row_title = "Deconvolution RNA",
  row_title_gp = gpar(fontsize = 14, fontface = "bold"),
  # cell_fun = cell_fun_maker(mat_PP_rna_pvalue),
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  # row_split = c(1,1,1,1,1,1,1,2),
  # column_split = c(1,1,1,1,1,1,1,2),
  show_heatmap_legend = TRUE,
  na_col = "grey90",
  # border = TRUE,
  rect_gp = gpar(col = "white", lwd = 2)
)
dev.off()
```