---
title: "HADACA3 Framework - Figure 2"
author: ""
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_document:
    self_contained: false
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
---


```{r, include=(FALSE)}
knitr::opts_chunk$set(collapse=TRUE, comment = "#>", fig.width=9, fig.height=6, eval=TRUE, echo=FALSE, fig.align = 'center')#, results="hide")
options(knitr.duplicate.label = "allow")
```

```{r params}
if (!exists("method")) {method = "TeamJ"}
if (!exists("type")) {type = "1"}
# source("datasets.R")

library(ggplot2)



datasets_long = c(
  "invitro", 
  "invivo",
  "insilicopseudobulk",
  "insilicodirichletNoDep",
  "insilicodirichletNoDep4CTsource",
  "insilicodirichletNoDep6CTsource",
  "insilicodirichletEMFA",
  "insilicodirichletEMFAImmuneLowProp",
  "insilicodirichletCopule"
)
datasets = c(
  "VITR", "VIVO", "SBN5", 
  "SDN5", "SDN4", "SDN6", 
  "SDE5", "SDEL", "SDC5"
)

# when you change the number of datasets you should change the number of heatmaps displayed (last lines of chunk Heatmap_pred)
```


```{r read_output}
ground_truth = lapply(datasets_long, function(x)
  readRDS(paste0("HADACA3_TeamJ_Results/01_groundtruth/groundtruth",type,"_",x,"_pdac.rds")))
names(ground_truth) = datasets

prediction = lapply(datasets, function(x)
  readRDS(paste0("HADACA3_TeamJ_Results/02_prediction/prediction",type,"_",method,"_",x,"_pdac.rds"))$prediction)
names(prediction) = datasets

time = lapply(datasets, function(x)
  readRDS(paste0("HADACA3_TeamJ_Results/02_prediction/prediction",type,"_",method,"_",x,"_pdac.rds"))$time)
names(time) = datasets

score = lapply(datasets, function(x)
  readRDS(paste0("HADACA3_TeamJ_Results/03_scores/score",type,"_",method,"_",x,"_pdac.rds")))
names(score) = datasets
```

# Visualisations of your method

## Table of scores for all datasets

```{r Display_scores}
score_df = data.frame(matrix(as.numeric(cbind(do.call(rbind,score),matrix(time,ncol=1))),nrow=length(datasets)))
rownames(score_df) = datasets
colnames(score_df) = c(names(score[[1]]), "time_exec")
knitr::kable(
  score_df, 
  # format = "html",
  table.attr = "style='overflow-x:auto; display:block; white-space:nowrap;'"
)
# kableExtra::scroll_box(score_table, width = "100%", height = "400px")
```

We have 9 different datasets: 

- `VITR` : *in vitro* mixtures of pure cell types
- `VIVO` : real *in vivo* bulk samples
- `SBN5` : *in silico simulation* of pseudo-bulk from simulated single-cell data, for 5 cell-types
- `SDN5` : *in silico simulation*, using a Dirichlet distribution, without correlation between features, for 5 cell-types
- `SDN4` : *in silico simulation*, using a Dirichlet distribution, without correlation between features, for 4 cell-types
- `SDN6` : *in silico simulation*, using a Dirichlet distribution, without correlation between features, for 6 cell-types
- `SDE5` : *in silico simulation*, using a Dirichlet distribution, with EMFA dependence, for 5 cell-types
- `SDEL` : *in silico simulation*, using a Dirichlet distribution, with EMFA dependence, for 5 cell-types including one with very low proportions
- `SDC5` : *in silico simulation*, using a Dirichlet distribution, with copula dependence, for 5 cell-types

For the `VIVO` dataset, we don't know the ground truth. However, we have a proxy for the relative proportions of cancer cell types, as measured on histological slides. Since we only have a partial ground truth, we can only compute correlation metrics on the cancer types.

```{r Heatmap_pred, eval=FALSE}
## Heatmap of predictions
for (dataset in seq_along(prediction)) {
  if (is.null(colnames(prediction[[dataset]]))) {
  colnames(prediction[[dataset]]) = paste0("Sample",
                             format(1:ncol(prediction[[dataset]]),
                                    digits = ceiling(log10(ncol(prediction[[dataset]])))))
  }
}

heatmaps = lapply(seq_along(prediction), function(x)
  ComplexHeatmap::Heatmap(
  prediction[[x]],
  col = circlize::colorRamp2(seq(0, 1, length.out = 32), rev(heat.colors(32))),
  column_title = datasets[x],
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  heatmap_legend_param = list(title = "Proportions"),
  column_names_rot = 45                                                         # HB: do we even show them ?
))
heatmaps[[1]] + heatmaps[[2]] + heatmaps[[3]]
#heatmaps[[4]] + heatmaps[[5]] + heatmaps[[6]]
#heatmaps[[7]] + heatmaps[[8]] + heatmaps[[9]]
```

## Boxplot of the absolute error

```{r abs_bias, warning=FALSE}
color_pallet = RColorBrewer::brewer.pal(9, "Set1")
names(color_pallet) = datasets
color_pallet = color_pallet[which(!is.na(names(color_pallet)))]

if (any(names(color_pallet) == "VIVO")) {
  color_pallet_abs_bias = color_pallet[-which(names(color_pallet) == "VIVO")]
} else {
  color_pallet_abs_bias = color_pallet
}

ground_truth_corrected = ground_truth
prediction_corrected = prediction
for (data in seq_along(datasets)) {
  if (nrow(prediction[[data]])==(nrow(ground_truth[[data]])-1)) { # we have 1 missing cell type in the ground truth
    if (all(rownames(prediction[[data]]) %in% rownames(ground_truth[[data]]))) {
      ground_truth_corrected[[data]] = ground_truth_corrected[[data]][rownames(prediction[[data]]),]
    }
    else {stop("Cell types names do not match")}
  }
  if (nrow(prediction[[data]])==(nrow(ground_truth[[data]])+1)) { # we have 1 missing cell type in the reference
    if (all(rownames(ground_truth[[data]]) %in% rownames(prediction[[data]]))) {
      extra_type = matrix(0,ncol=ncol(ground_truth[[data]]),nrow=1)
      rownames(extra_type) = setdiff(rownames(prediction[[data]]),rownames(ground_truth[[data]]))
      ground_truth_corrected[[data]] = rbind(ground_truth_corrected[[data]],extra_type)
      prediction_corrected[[data]] = prediction_corrected[[data]][rownames(ground_truth_corrected[[data]]),]
    }
    else {stop("Cell types names do not match")}
  }
  prediction_corrected[[data]] = prediction_corrected[[data]][rownames(ground_truth_corrected[[data]]),]
}

cell_types = rownames(prediction[[1]])
abs_bias = lapply(seq_along(prediction), function(x) abs(as.matrix(prediction_corrected[[x]][rownames(ground_truth_corrected[[x]]),]) - as.matrix(ground_truth_corrected[[x]][rownames(ground_truth_corrected[[x]]),])))
abs_bias = data.frame(
  Error = unlist(lapply(abs_bias,c)),
  CellType = unlist(lapply(abs_bias, function(x) rep(rownames(x),ncol(x)))),
  Dataset = unlist(lapply(seq_along(abs_bias), function(x) rep(datasets[x],ncol(abs_bias[[x]])*nrow(abs_bias[[x]]))))
)
abs_bias$Dataset = factor(abs_bias$Dataset, levels = datasets)
abs_bias[abs_bias$Dataset == "VIVO", "Error"] = NA_real_

# boxplots = lapply(cell_types, function(CT)
#   ggplot(abs_bias, aes(x=Dataset, y=Error, color=Dataset)) +
#     geom_boxplot() +
#     geom_point() +
#     xlab("") )
#     #see::theme_modern(axis.text.angle = 40, axis.text.size = 5))
# #ggpubr::ggarrange(plotlist = boxplots, common.legend = T, labels = cell_types)
# #boxplots[[1]]

boxplots = ggplot(abs_bias, aes(x=CellType, y=Error, fill=Dataset)) +
  geom_boxplot(position = position_dodge2(padding = 0.2)) + # 
  labs(
    title = paste("Absolute bias among each cell type by your method on each dataset"),
    x = "Cell types",
    y = "Absolut bias"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom"
  ) +
  scale_fill_manual(
    values = color_pallet_abs_bias
  )
plotly::layout(
  plotly::ggplotly(boxplots), 
  boxmode = "group",
  legend = list(
    # x = 0,
    # y = -0.3,
    # xanchor = 'left',
    # yanchor = 'bottom',
    orientation = 'h')
)
```

Boxplot of absolute error between estimated proportions $\hat{\beta}$ and ground truth proportions $\beta$ for each cell type. For the sample $j$ and the cell type $i$,  $\text{Absolut error}_{i,j} = |\hat{\beta}_{i,j} - \beta_{i,j}|$

## Boxplot of each metric

```{r scores_samples_all_metrics, warning=FALSE}
correlationP_1col = function(Ai_real, Ai_pred) {
  # Ai_pred = Ai_pred[names(Ai_real)]
  return(cor(Ai_real, Ai_pred, method = "pearson"))
}
correlationS_1col = function(Ai_real, Ai_pred) {
  # Ai_pred = Ai_pred[names(Ai_real)]
  return(cor(Ai_real, Ai_pred, method = "spearman"))
}
eval_Aitchison_1col = function(Ai_real, Ai_pred, min = 1e-9) {
  # Ai_pred = Ai_pred[names(Ai_real)]
  # Aitchison dist doesn't like 0 
  Ai_real[Ai_real < min] = min
  Ai_pred[Ai_pred < min] = min
  return(coda.base::dist(rbind(Ai_real, Ai_pred), method = "aitchison")[1])
}
eval_Aitchison = function(A_real, A_pred, min = 1e-9) {
  # A_pred = A_pred[rownames(A_real),]
  # Aitchison dist doesn't like 0 
  A_real[A_real < min] = min
  A_pred[A_pred < min] = min
  res = c()
  for (i in seq(ncol(A_real))) {
    res[i] = coda.base::dist(rbind(t(A_real[,i]), t(A_pred[,i])), method = "aitchison")[1]
  }
  res = res[!is.na(res)]
  return(mean(res))
}
eval_RMSE = function(A_real, A_pred) {
  return(sqrt(mean((A_real - A_pred)^2)))
}
eval_MAE = function (A_real, A_pred){
  return(mean(abs(A_real - A_pred)))
}




df_metric = do.call(
  what = rbind,
  args = lapply(seq_along(datasets), function(data) {
    
    stopifnot(identical(rownames(prediction_corrected[[data]]), rownames(ground_truth_corrected[[data]])))
    prediction_corrected[[data]][rownames(ground_truth_corrected[[data]]),]

    data.frame(
      "RMSE" = sapply(seq(ncol(ground_truth_corrected[[data]])), function(sample) {
        eval_RMSE(ground_truth_corrected[[data]][,sample], prediction_corrected[[data]][,sample])
      }),
      "MAE" = sapply(seq(ncol(ground_truth_corrected[[data]])), function(sample) {
        eval_MAE(ground_truth_corrected[[data]][,sample], prediction_corrected[[data]][,sample])
      }),
      "Aitchison" = sapply(seq(ncol(ground_truth_corrected[[data]])), function(sample) {
        eval_Aitchison_1col(ground_truth_corrected[[data]][,sample], prediction_corrected[[data]][,sample])
      }),
      "Pearson" = sapply(seq(ncol(ground_truth_corrected[[data]])), function(sample) {
        correlationP_1col(ground_truth_corrected[[data]][,sample], prediction_corrected[[data]][,sample])
      }),
      "Spearman" = sapply(seq(ncol(ground_truth_corrected[[data]])), function(sample) {
        correlationS_1col(ground_truth_corrected[[data]][,sample], prediction_corrected[[data]][,sample])
      }),
      "Dataset" = datasets[data]
    )
  })
)
  
df_metric = tidyr::pivot_longer(
  data = df_metric,
  cols = c("RMSE", "MAE", "Aitchison", "Pearson", "Spearman"),
  names_to = "metric_name",
  values_to = "metric_score"
) 
df_metric$Dataset = factor(df_metric$Dataset, levels = datasets)
df_metric$metric_name = factor(df_metric$metric_name, levels = c("RMSE", "MAE", "Aitchison", "Pearson", "Spearman"))
df_metric[(df_metric$Dataset == "VIVO") & (df_metric$metric_name %in% c("RMSE", "MAE", "Aitchison")), "metric_score"] = NA_real_

# boxplots_metric = lapply(c("RMSE","MAE","Aitchison","Pearson","Spearman"), function(metric)
#   ggplot(df_metric, aes(x=Dataset, y=get(metric), color=Dataset)) +
#     geom_boxplot() +
#     geom_point() +
#     xlab("") +
#     ylab(metric) )
#     #see::theme_modern(axis.text.angle = 40, axis.text.size = 5))
# #ggpubr::ggarrange(plotlist = boxplots_metric, common.legend = T)
# boxplots_metric[[1]]
# boxplots_metric[[2]]

boxplots_metric = ggplot(df_metric, aes(x = Dataset, y = metric_score, fill = Dataset)) +
  facet_wrap( ~ metric_name, scales = "free") +
  geom_boxplot(position=position_dodge2(padding = 0.2)) +
  labs(
    title = paste("Samples' metrics results of your method on all datasets"),
    x = "Datasets",
    y = "Scores"
  ) +
  theme_light() +
  theme(
    plot.title = element_text(hjust = 0.5),
    legend.position = "bottom",
    axis.text.x = element_blank() # element_text(angle = 45, hjust = 1)
  ) +
  scale_fill_manual(
    values = color_pallet
  )
pp = plotly::ggplotly(boxplots_metric, width = 900, height = 650)
# pp
pp = plotly::layout(pp, legend = list(orientation = 'h'))
# pp
invisible(
  lapply(seq_along(pp$x$layout$shapes), \(i){ # ça ne regle pas entierement le pb, ils sont toujours sur les plots du dessus
  
  # if a strip shape (using the strip color), reduce the height of the strip box
  if(isTRUE(pp$x$layout$shapes[[i]]$fillcolor == "rgba(179,179,179,1)")) {
   pp$x$layout$shapes[[i]]$y1 <<- 18    # base at 23.379
  }
  # there are less annotations than shapes
  if(i <= length(pp$x$layout$annotations)) { 
    # reduce the font size of the strip text
    if(any(pp$x$layout$annotations[[i]]$text %in% unique(df_metric$metric_name))) {
      pp$x$layout$annotations[[i]]$font$size <<- 11   # base at 11.6895
    }
  }
})
)
pp

# plotly::layout(
#   pp,
#   legend = list(
#     # x = 0,
#     # y = 0,#-0.2,
#     # xanchor = 'left',
#     # yanchor = 'right',
#     orientation = 'h'),
#   # ggplotly make strip are on top of upper plot (which is not nice)
#   # the simplest way (but ugly) is to add space for each strip position
#   # yaxis = list(range = c(-0.05, max(df_metric[df_metric$metric_name == "RMSE", "metric_score"], na.rm = TRUE)))#,
#   # yaxis2 = list(range = c(-0.5, 1)),
#   # yaxis3 = list(range = c(-0.5, 1))
# )
```


Boxplot of 15 or 30 samples for five different metrics : between estimated proportions and ground truth proportions for each cell type

- Root Mean Squared Error: for the sample $j$, $\text{RMSE}_j = \sqrt{\frac{\sum_{i=1}^{N} \big(\hat{\beta}_{i,j}-\beta_{i,j}\big)^2}{N}}$.
- Mean Absolut Error: for the sample $j$, $\text{MAE}_j = \frac{\sum_{i=1}^{N} \big|\hat{\beta}_{i,j}-\beta_{i,j}\big|}{N}$.
- Aitchison distance: this distance use a isomorphism to transform a vector of proportions, true and estimated vectors for each samples, from the unity simplex to the real space (see https://en.wikipedia.org/wiki/Compositional_data#Aitchison_geometry). Then, an Euclidean distance is calculated between the transformed estimated vector of proportions and the transformed ground truth vector of proportions.
- Pearson correlation: for the sample $j$, the Pearson correlation is calculated between the estimated vector of proportions $\hat{\beta}_j$ and the ground truth vector of proportions $\beta_j$.
- Spearman correlation: for the sample $j$, the Spearman correlation is calculated between the estimated vector of proportions $\hat{\beta}_j$ and the ground truth vector of proportions $\beta_j$.



## Scatter plots between truth and prediction

```{r scatter_plot, warning=FALSE}
df_scatter = do.call(
  what = rbind,
  args = lapply(seq_along(datasets), function(data) { # we remove in vivo
    
    prediction_data = t(prediction_corrected[[data]])
    prediction_data = tidyr::pivot_longer(
      as.data.frame(prediction_data),
      cols = colnames(prediction_data),
      names_to = "cell_type",
      values_to = "pred_prop"
    )
    
    ground_truth_data = t(ground_truth_corrected[[data]])
    ground_truth_data = tidyr::pivot_longer(
      as.data.frame(ground_truth_data),
      cols = colnames(ground_truth_data),
      names_to = "cell_type",
      values_to = "true_prop"
    )
    
    prop_data = cbind(prediction_data, ground_truth_data[,"true_prop"], Dataset = datasets[data])
  })
)
df_scatter$Dataset = factor(df_scatter$Dataset, levels = datasets)
df_scatter$cell_type = factor(df_scatter$cell_type, levels = rownames(ground_truth[[1]]))


scatter_prop = ggplot(df_scatter) +
  facet_wrap( ~ Dataset) +
  geom_abline(intercept = 0, slope = 1, linetype = 3, linewidth = 0.25) +
  geom_point(aes(x = true_prop, y = pred_prop, col = cell_type)) +
  coord_cartesian(
    xlim = c(0,1),
    ylim = c(0,1)
  ) +
  labs(
    x = "Ground truth proportion",
    y = "Prediction",
    title = paste("Proportions predicted by your method vs true ones for all datasets")
  ) +
  scale_x_continuous(
    labels = scales::percent
  ) +
  scale_y_continuous(
    labels = scales::percent
  ) +
  scale_color_discrete(
    name = "Cell type"
  ) +
  theme_light() +
  theme(plot.title = element_text(hjust = 0.5))

pp = plotly::ggplotly(scatter_prop, width = 900, height = 1000)
# pp
pp = plotly::layout(pp, legend = list(orientation = 'h'))
# pp
invisible(
  lapply(seq_along(pp$x$layout$shapes), \(i){ # ça ne regle pas entierement le pb, ils sont toujours sur les plots du dessus
  
  # if a strip shape (using the strip color), reduce the height of the strip box
  if(isTRUE(pp$x$layout$shapes[[i]]$fillcolor == "rgba(179,179,179,1)")) {
   pp$x$layout$shapes[[i]]$y1 <<- 18    # base at 23.379
  }
  # there are less annotations than shapes
  if(i <= length(pp$x$layout$annotations)) { 
    # reduce the font size of the strip text
    if(any(pp$x$layout$annotations[[i]]$text %in% unique(df_metric$metric_name))) {
      pp$x$layout$annotations[[i]]$font$size <<- 11   # base at 11.6895
    }
  }
})
)
pp
```

## Radar charts

```{r scores_global_radarchart, fig.width=20, fig.height=11.25, warning=FALSE, message=FALSE, results='hide'}
if (length(datasets) == 9) {
  par(mfrow = c(3,3))
} else if (length(datasets) > 1) {
  par(mfrow = c(length(datasets) %/% 3, 3))
}

for (data in seq_along(datasets)) {
  ground_truth_data = ground_truth[[data]]
  
  fake_worst_pred = apply(ground_truth_data, 2, function(prop) { 
    tmp = rep(1e-9, length(prop))
    tmp[which.min(prop)] = 1
    return(tmp)
  })
  rownames(fake_worst_pred) = rownames(ground_truth_data)
  
  worst_RMSE = eval_RMSE(ground_truth_data, fake_worst_pred)
  worst_MAE = eval_MAE(ground_truth_data, fake_worst_pred)
  worst_Aitchison = eval_Aitchison(ground_truth_data, fake_worst_pred)
  
  
  score_data = score[[data]]
  
  # We don't take the global score for the radar chart since it's a proxy of the area
  df_radarplot = as.data.frame(rbind(
    # rmse,       mae,       aitchison,        pearson_tot, pearson_col, pearson_row, spearman_tot, spearman_col, spearman_row
    c(0,          0,         0,                1,           1,           1,           1,            1,            1), # the max/best value for each var
    c(worst_RMSE, worst_MAE, worst_Aitchison, -1,          -1,          -1,          -1,           -1,           -1), # the min/worst value for each var
    score_data[-1]
  ))
  
  fmsb::radarchart(
    df_radarplot, 
    axistype=2, # add the max/best value of each axe (but I don't know how to add the min/worst)
    
    title = paste("Global metrics scores for your method\non dataset", datasets[data]),
    #custom polygon
    pcol=rgb(0.2,0.5,0.5,0.9), pfcol=rgb(0.2,0.5,0.5,0.5), plwd=4,
   
    #custom the grid
    cglcol="grey", cglty=1, axislabcol="grey50", cglwd=0.8,
   
    #custom labels
    vlcex=0.8
  )
  
  # Comment HB: the visualization by a radar/spider plot is intrinsically linked to the aggregated score for defining the scope (min and max) for each variable.
}
par(mfrow = c(1,1))
```

# Session Information

```{r, results="verbatim"}
sessionInfo()
```
